---
title: "01 — QAQC of HOBO conductivity data from nutrient addition slugs"
subtitle: "Karst streams, dry season 2023"
author: "Kauan Fonseca"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: show
    theme: flatly
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6
)

library(readxl)
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
library(lubridate)
library(hms)
library(ggplot2)
library(plotly)
library(here)
```

# Scope

This document covers **instrument-level QAQC only**: import of the raw
HOBO conductivity exports, harmonization of headers, recalculation of
specific conductance, anchoring of each series to its tracer addition,
interactive selection of the slug window, and export of a clean
high-resolution conservative-tracer series.

It deliberately stops before any cross-instrument work. Merging with the
nutrient grab samples, reconciling the handheld probe against the logger,
correcting SRP for CaCO₃ co-precipitation, and building the analysis
tables are handled in `02_integration.Rmd`. Visual review of the final
breakthrough curves lives in `03_btc_review.Rmd`.

**Output of this document:** `data_derived/btc_conservative.csv` (+ RDS)
and `data_derived/events_hobo.csv`.

# Sampling protocol

Specific conductance (SpC), full-range conductivity, and water
temperature were recorded continuously (5–10 s intervals) with HOBO
U24-001 conductivity loggers deployed during combined NaCl + nutrient
slug releases. Loggers were deployed at a single downstream station for
one-station slugs, or at paired upstream/downstream stations. Some
streams were sampled on a single day; others include multiple sampling
dates or repeated releases within the same day.

All timestamps in this pipeline are standardised to Brasília time
(`America/Sao_Paulo`, UTC−3), which is the logger's native clock.
Field notebooks recorded local Mato Grosso time (`America/Cuiaba`,
UTC−4); conversion happens at the point of import and is documented
where it occurs.

---

# 1. Import raw HOBO exports

Raw HOBO data were exported per stream and sampling day as individual
`.xlsx` files. Each file is imported and stacked into a single raw data
frame, retaining the source file name as a traceability field.

```{r list-files}
dir_hobo <- here("data", "hobo_raw")

files_all <- list.files(dir_hobo, pattern = "\\.xlsx$", full.names = TRUE)

tibble(file = basename(files_all))
```

## 1.1 Excluded deployments

```{r exclude-files}
# Files excluded from the pipeline, with the reason recorded so the
# exclusion is auditable rather than silent.
exclusions <- tribble(
  ~file,                    ~reason,
  "SpC_CD_061023.xlsx",     "Export contained columns from multiple logger deployment sessions (light intensity and dissolved oxygen unrelated to the slug), preventing extraction of a single SpC/temperature series. CD 2023-10-09 is used as the valid record for this stream."
)

files <- files_all[!basename(files_all) %in% exclusions$file]

tibble(
  excluded_file  = exclusions$file,
  found_in_data  = exclusions$file %in% basename(files_all)
)

length(files_all); length(files)
```

> **QAQC note.** `SpC_CD_061023.xlsx` was removed from `data/hobo_raw/`
> during project reorganisation, so the filter above is currently a
> no-op. It is retained as an explicit, documented rule: if the file is
> ever restored to the folder it will still be excluded, and the reason
> travels with the code.

```{r import-all}
hobo_raw_list <- files %>%
  set_names(basename(.)) %>%
  map(~ read_excel(.x, sheet = 1))

hobo_raw_all <- bind_rows(hobo_raw_list, .id = "source_file")

nrow(hobo_raw_all)
```

# 2. Inspect raw structure

```{r inspect-columns}
map(hobo_raw_list, names)
```

```{r inspect-dims}
imap_dfr(hobo_raw_list, ~ tibble(
  source_file = .y, n_rows = nrow(.x), n_cols = ncol(.x)
))
```

# 3. Standardise column names

Header text varies across files because of the HOBOware interface
language (Portuguese vs English) and because the logger serial number is
embedded in the header (e.g. `#20835561`). A regex mapping brings every
file to a common schema. The serial is extracted *before* renaming, so
it is preserved as a metadata field.

```{r fn-standardize}
extract_logger_serial <- function(df) {
  serial <- names(df) %>% str_extract("#\\d+") %>% na.omit() %>% unique()

  if (length(serial) == 0) return(NA_character_)
  if (length(serial) > 1) {
    warning("Multiple logger serials in one file: ", paste(serial, collapse = ", "))
    return(paste(str_remove(serial, "#"), collapse = "; "))
  }
  str_remove(serial, "#")
}

standardize_hobo_names <- function(df) {
  names(df) <- names(df) %>%
    str_replace("(?i)^date$", "date_raw") %>%
    str_replace("(?i)^time$", "time_raw") %>%
    str_replace("(?i)^(Full Range|Gama Completa).*", "full_range_uscm") %>%
    str_replace("(?i)^Temp\\.?\\s*\\(°C\\).*", "temp_c") %>%
    str_replace("(?i)^Specific Conductance.*", "spc_hoboware_uscm") %>%
    str_replace("(?i)^(Coupler Detached|Acoplador Desligado).*", "coupler_detached") %>%
    str_replace("(?i)^(Coupler Attached|Acoplador Ligado).*", "coupler_attached") %>%
    str_replace("(?i)^(Host Connected|Anfitrião Ligado).*", "host_connected") %>%
    str_replace("(?i)^(Stopped|Parado)$", "stopped") %>%
    str_replace("(?i)^(End Of File|Fim Do Ficheiro).*", "end_of_file")
  df
}

logger_serials <- map_chr(hobo_raw_list, extract_logger_serial)
hobo_std_list  <- map(hobo_raw_list, standardize_hobo_names)

tibble(source_file = names(logger_serials), logger_serial = logger_serials)
```

```{r check-schema}
# every file must expose the three columns the pipeline depends on
map_dfr(hobo_std_list, ~ tibble(
  has_datetime  = all(c("date_raw", "time_raw") %in% names(.x)),
  has_fullrange = "full_range_uscm" %in% names(.x),
  has_temp      = "temp_c" %in% names(.x)
), .id = "source_file")
```

# 4. File-level metadata

```{r extract-metadata}
file_metadata <- tibble(source_file = names(hobo_std_list)) %>%
  mutate(
    stream = str_extract(source_file, "(?<=SpC_)[A-Z]{2}"),
    station = case_when(
      str_detect(source_file, "upstream")  ~ "upstream",
      # "dowstream" (sic) is the typo baked into the original raw filenames;
      # matched alongside the correctly-spelled form so a new, correctly
      # named file (e.g. SR 2023-10-09) is not silently left as NA.
      str_detect(source_file, "dow(n)?stream") ~ "downstream",
      TRUE ~ NA_character_
    ),
    date_str      = str_extract(source_file, "\\d{6}"),
    date          = dmy(date_str),
    logger_serial = logger_serials[source_file]
  ) %>%
  select(-date_str)

file_metadata
```

```{r fix-cb-date}
# SpC_CB_100823.xlsx: filename follows ddmmyy, but the day and month
# digits were transposed when the file was named. The true event date is
# 2023-10-08, confirmed against the field notebook. CB was sampled on two
# separate dates (2023-09-07 and 2023-10-08); both are valid events.
file_metadata <- file_metadata %>%
  mutate(date = if_else(source_file == "SpC_CB_100823.xlsx",
                        as.Date("2023-10-08"), date))

file_metadata %>% filter(stream == "CB")
```

```{r stack-standardized}
hobo_std_all <- bind_rows(hobo_std_list, .id = "source_file") %>%
  left_join(file_metadata, by = "source_file")

glimpse(hobo_std_all)
```

# 5. Datetime and specific conductance

```{r build-datetime}
# HOBOware exports Date and Time as two full datetimes carrying the same
# calendar date; recombine into one clean POSIXct in Brasília time.
hobo_std_all <- hobo_std_all %>%
  mutate(
    datetime = as.POSIXct(
      paste(format(as.Date(date_raw), "%Y-%m-%d"), format(time_raw, "%H:%M:%S")),
      format = "%Y-%m-%d %H:%M:%S", tz = "America/Sao_Paulo"
    )
  ) %>%
  arrange(source_file, datetime)

range(hobo_std_all$datetime, na.rm = TRUE)
sum(is.na(hobo_std_all$datetime))
```

```{r recalc-spc}
# Temperature compensation: 2 % per °C, referenced to 25 °C
# SpC = full_range / (1 + 0.02 * (temp_c - 25))
hobo_std_all <- hobo_std_all %>%
  mutate(spc_uscm = full_range_uscm / (1 + 0.02 * (temp_c - 25)))
```

```{r compare-spc-sources}
# how far does the recalculated value sit from HOBOware's own?
hobo_std_all %>%
  filter(!is.na(spc_hoboware_uscm)) %>%
  group_by(source_file) %>%
  summarise(
    n              = n(),
    median_diff    = median(spc_hoboware_uscm - spc_uscm, na.rm = TRUE),
    max_abs_diff   = max(abs(spc_hoboware_uscm - spc_uscm), na.rm = TRUE),
    hoboware_min   = min(spc_hoboware_uscm, na.rm = TRUE),
    .groups = "drop"
  )
```

> **QAQC note.** HOBOware's pre-calculated Specific Conductance
> (`spc_hoboware_uscm`) was unreliable for at least one deployment
> (RS, 2023-10-11), flatlining at 1 µS/cm through the peak of the slug
> despite plausible `full_range_uscm` and `temp_c`. Rather than repair
> one file, `spc_uscm` is recalculated for **every** event directly from
> `full_range_uscm` and `temp_c` using the standard 2 %/°C compensation
> referenced to 25 °C. The HOBOware column is retained only for the
> diagnostic comparison above and is never used downstream.

# 6. Event metadata

Tracer masses, reach geometry and release times come from the field
sheet (`slugs_kauan_clean.csv`).

```{r read-field-sheet}
field_sheet <- read_csv(here("data", "field", "slugs_kauan_clean.csv"),
                        show_col_types = FALSE)

field_sheet %>% distinct(stream, date) %>% arrange(stream, date)
```

```{r addition-times}
# Release times were written down in Mato Grosso local time (UTC-4) and
# are converted here, at the point of import, to Brasília time (UTC-3),
# the logger's clock and the standard for the rest of the pipeline.
addition_times <- field_sheet %>%
  filter(time_phase == "release") %>%
  group_by(stream, date) %>%
  summarise(addition_time_local = first(time_hms), .groups = "drop") %>%
  mutate(
    addition_datetime = with_tz(
      as.POSIXct(paste(format(date, "%Y-%m-%d"), addition_time_local),
                 format = "%Y-%m-%d %H:%M:%S", tz = "America/Cuiaba"),
      tzone = "America/Sao_Paulo"
    ),
    # re-extract date and time already in Brasília time; covers the rare
    # case of a release close to midnight rolling the calendar date
    addition_time = hms::as_hms(addition_datetime),
    date          = as.Date(addition_datetime)
  ) %>%
  select(stream, date, addition_time, addition_datetime)

event_metadata <- field_sheet %>%
  distinct(stream, date, nacl_mass_g, amonium_added_g, phophate_added_g,
           reach_length_m, reach_mean_width_m) %>%
  left_join(addition_times, by = c("stream", "date"))

# must be empty: one metadata row per stream-date
event_metadata %>% count(stream, date) %>% filter(n > 1)

event_metadata
```

```{r manual-events}
# SA 2023-10-10 was captured by the logger but never entered in the
# field sheet; transcribed here from the field notebook.
manual_events <- tribble(
  ~stream, ~date,                 ~addition_time_local,    ~nacl_mass_g, ~amonium_added_g, ~phophate_added_g, ~reach_length_m, ~reach_mean_width_m,
  "SA",    as.Date("2023-10-10"), hms::as_hms("12:45:00"), 3139,         155,              298,               200,             10
) %>%
  mutate(
    addition_datetime = with_tz(
      as.POSIXct(paste(format(date, "%Y-%m-%d"), addition_time_local),
                 format = "%Y-%m-%d %H:%M:%S", tz = "America/Cuiaba"),
      tzone = "America/Sao_Paulo"
    ),
    addition_time = hms::as_hms(addition_datetime),
    date          = as.Date(addition_datetime)
  ) %>%
  select(-addition_time_local)

event_metadata <- bind_rows(event_metadata, manual_events) %>%
  arrange(stream, date)

event_metadata
```

> **QAQC note — SR 2023-10-09 (downstream station).** A second SR
> deployment, two days before the nutrient-addition event already in the
> field sheet (SR 2023-10-11). This one was a discharge-characterisation
> slug only (no nutrients added); the field sheet has no row for it, so
> it is added manually below, mirroring the SA 2023-10-10 case above.
> `reach_length_m = 72` is not a new value — it is the same downstream
> distance already on record for SR 2023-10-11, confirming the two
> deployments share a station. `nacl_mass_g = 3101` is from Kauan's field
> summary for this slug.
>
> The release time was recovered, not inferred: an earlier export of this
> same deployment (`data/field/SR_discharge_091023_downstream.xlsx`,
> before Kauan re-exported it with temperature added, which is what is
> imported above) carried its own `time since release` column, zeroed at
> **2023-10-09 17:33:10** Brasília time — inside the recalled 17:30-18:30
> window, and directly on the logger's own clock, so no Cuiabá
> conversion applies. That anchor puts the SpC peak (18:15:50) at
> t = 2560 s, matching the peak time this project already had on record
> for this event before the file was re-exported. It is adopted here.
> **QAQC note — SR 2023-10-09 upstream station.** The same logger serial
> (`#20835564`) that recorded this slug also runs continuously for daily
> SpC/O₂ metabolism monitoring — that dual purpose is the reason its raw
> file carries an unrelated sheet name and spans a full ~8 h deployment
> rather than a tight slug window; it is not a different, unrelated
> event. `reach_length_m = 52` is Kauan's approximate field distance
> (unlike the 72 m downstream value, this one is not cross-confirmed
> against an independent record, so it carries more uncertainty).
> `addition_datetime` is shared with the downstream station rather than
> read from this file's own `time since release` column: both stations
> sample the same single injection, so there is only one true release
> instant, and this file's own internal zero-point (16:34:00) does not
> agree with it (see discussion in `02`). Width is not separately known
> for this station and reuses the stream-level 2 m.
>
> `event_metadata` is keyed by `stream + date` only (Section 9 is what
> resolves the `station` split into two distinct `event_id`s), so it
> cannot carry two different `reach_length_m` values for the same
> stream-date the way it would need to for 72 m (downstream) vs 52 m
> (upstream) — the existing RA 2023-10-05 downstream/upstream pair has
> this same limitation and both already share one `reach_length_m` in
> `events_hobo`. The row below carries the downstream (72 m) value,
> consistent with that precedent; the upstream-specific 52 m is applied
> as a per-`event_id` override in `02`, where it is used (only the
> logger integral, not distance, drives `discharge_Ls_logger`, but
> distance does drive `water_velocity_ms`).
```{r manual-events-sr}
manual_events_sr <- tribble(
  ~stream, ~date,                 ~nacl_mass_g, ~amonium_added_g, ~phophate_added_g, ~reach_length_m, ~reach_mean_width_m,
  "SR",    as.Date("2023-10-09"), 3101,         0,                0,                 72,              2
) %>%
  mutate(
    addition_datetime = as.POSIXct("2023-10-09 17:33:10", tz = "America/Sao_Paulo"),
    addition_time     = hms::as_hms(addition_datetime)
  )

event_metadata <- bind_rows(event_metadata, manual_events_sr) %>%
  arrange(stream, date)

event_metadata %>% filter(stream == "SR")
```

```{r coverage-check}
# which logged events have no field-sheet metadata, and vice versa?
file_metadata %>%
  distinct(stream, date) %>%
  mutate(has_logger = TRUE) %>%
  full_join(event_metadata %>% distinct(stream, date) %>% mutate(has_metadata = TRUE),
            by = c("stream", "date")) %>%
  arrange(stream, date)
```

## 6.1 Background reference from the handheld probe

```{r background-reference}
background_reference <- field_sheet %>%
  filter(time_phase == "release") %>%
  distinct(stream, date, backg_spc_uScm) %>%
  rename(background_spc_probe = backg_spc_uScm) %>%
  bind_rows(tibble(stream = "SA", date = as.Date("2023-10-10"),
                   background_spc_probe = 5.1)) %>%
  arrange(stream, date)

background_reference
```

> **QAQC note.** `background_spc_probe` is **not** derived from the
> logger. It is the specific conductance read with the handheld probe
> (Hanna HI12883 or YSI) at the moment of release, kept here strictly as
> an independent reference value. It is useful for detecting drift
> between the two instruments but is not assumed to be numerically
> identical to the logger's own baseline. The probe-vs-logger
> reconciliation is carried out in `02_integration.Rmd`.

```{r join-meta}
hobo_with_meta <- hobo_std_all %>%
  left_join(event_metadata,       by = c("stream", "date")) %>%
  left_join(background_reference, by = c("stream", "date"))

hobo_with_meta %>% count(source_file, stream, date, station)
```

# 7. Tracer addition anchors

Most events had a single combined NaCl + nutrient release. RA on
2023-09-06 is the exception: phosphorus and nitrogen were added
separately, at 14:30 and 16:45, with the logger deployed from 13:39 to
18:35. **These two times were already recorded in Brasília time and must
not pass through the Cuiabá conversion applied to every other event.**

```{r addition-anchor}
addition_anchor_standard <- event_metadata %>%
  filter(!(stream == "RA" & date == as.Date("2023-09-06"))) %>%
  transmute(stream, date, slug_label = NA_character_,
            addition_datetime,
            label = if_else(coalesce(amonium_added_g, 0) == 0 & coalesce(phophate_added_g, 0) == 0,
                            "NaCl-only discharge slug", "NaCl + nutrient release"))

addition_anchor_ra <- tribble(
  ~stream, ~date,                 ~slug_label, ~addition_datetime,                                        ~label,
  "RA",    as.Date("2023-09-06"), "P",  as.POSIXct("2023-09-06 14:30:00", tz = "America/Sao_Paulo"), "P addition",
  "RA",    as.Date("2023-09-06"), "N",  as.POSIXct("2023-09-06 16:45:00", tz = "America/Sao_Paulo"), "N addition"
)

addition_anchor <- bind_rows(addition_anchor_standard, addition_anchor_ra) %>%
  arrange(stream, date, slug_label)

addition_anchor
```

> **QAQC note.** `event_metadata` holds one row per stream-date, which
> is correct for reach geometry and total masses but wrong as a time
> anchor for RA 2023-09-06, where two slugs share a calendar date. The
> anchor table above is keyed on `stream + date + slug_label` and is the
> **only** source of `addition_datetime` used to compute elapsed time.
> Any `addition_datetime` inherited through the event-level join is
> dropped before the anchor is applied (Section 9).

## 7.1 Full raw series review, before trimming

```{r plot-full-series}
p_full <- hobo_with_meta %>%
  ggplot(aes(x = datetime, y = spc_uscm)) +
  geom_line(linewidth = 0.3) +
  geom_vline(
    data = addition_anchor %>%
      inner_join(file_metadata %>% select(source_file, stream, date),
                 by = c("stream", "date")),
    aes(xintercept = addition_datetime),
    colour = "darkgreen", linetype = "dashed"
  ) +
  facet_wrap(~ source_file, scales = "free") +
  labs(x = NULL, y = "SpC (µS/cm)",
       title = "Full raw HOBO series per deployment",
       subtitle = "Dashed green = tracer addition") +
  theme_minimal(base_size = 10)

ggplotly(p_full)
```

# 8. Interactive slug window selection

The slug window (release to return to background, or to logger
retrieval) was selected per event with a dedicated Shiny application,
`R/app_slug_interval.R`, which brushes over the high-resolution series
with zoom and hover inspection of exact SpC/datetime values. It writes
one trimmed CSV per event to `data/slug_trimmed/` and appends every
selection to `data/slug_trimmed_log.csv`.

```{r save-for-shiny}
saveRDS(hobo_with_meta, here("data", "hobo_with_meta.rds"))
```

> Run `R/app_slug_interval.R` at this point if the trimmed files need to
> be regenerated. They are version-controlled, so the rest of this
> document knits without re-running the app.

## 8.1 SR 2023-10-09 (downstream): window built without the Shiny app

This single new event was not run through `app_slug_interval.R` — the
interactive app cannot run inside this session. It did not need real
trimming judgement anyway: the raw export already opens on a flat
pre-release baseline and closes on a flat post-recession baseline, so
the whole file *is* the slug window. Only the trailing non-numeric sentinel
rows the logger appends past the last measurement (a stray `"Logged"`
text token, no `full_range_uscm`) are dropped. The result is written
directly to `data/slug_trimmed/`, in the same column layout as every
other trimmed file, so Section 9 onward treats it identically to a
Shiny-produced window; the substitution is logged as its own row in
`data/slug_trimmed_log.csv` rather than silently blended in.

```{r sr-manual-trim}
sr_window <- hobo_with_meta %>%
  filter(stream == "SR", station == "downstream", date == as.Date("2023-10-09")) %>%
  filter(!is.na(full_range_uscm), !is.na(temp_c)) %>%
  arrange(datetime) %>%
  # the last logged row (19:18:20) jumps from 501 to 539 uS/cm in a single
  # 10 s step, right after five consecutive readings at the pre-release
  # baseline (i.e. the tracer had already fully returned to background) —
  # a sensor-retrieval artefact, not water. An earlier export of this same
  # file that still carried a precomputed NaCl column showed the identical
  # spike at this exact timestamp, so it is not an artefact of this
  # re-import. Dropped, with the reason kept here rather than silently.
  filter(datetime != max(datetime)) %>%
  transmute(
    source_file, date_raw, time_raw, full_range_uscm, temp_c,
    spc_calculated_uscm = spc_uscm,
    coupler_detached = NA_real_, coupler_attached = NA_real_,
    host_connected = NA_real_, stopped = NA_real_, end_of_file = NA_real_,
    stream, station, date, logger_serial, datetime, spc_uscm,
    nacl_mass_g, amonium_added_g, phophate_added_g,
    reach_length_m, reach_mean_width_m,
    addition_time, addition_datetime,
    background_spc = background_spc_probe
  )

nrow(sr_window); range(sr_window$datetime)

write_csv(sr_window, here("data", "slug_trimmed", "slug_SR_20231009_downstream.csv"))

sr_log_row <- tibble(
  source_file   = "SpC_SR_091023_downstream.xlsx",
  stream        = "SR",
  date          = as.Date("2023-10-09"),
  station       = "downstream",
  start_datetime = min(sr_window$datetime),
  end_datetime   = max(sr_window$datetime),
  n_obs          = nrow(sr_window),
  output_file    = "slug_SR_20231009_downstream.csv",
  label          = "built directly from the raw export (Shiny app not run this session); see Sec. 8.1"
)

log_path <- here("data", "slug_trimmed_log.csv")
existing_log <- read_csv(log_path, show_col_types = FALSE)
if (!sr_log_row$output_file %in% existing_log$output_file) {
  write_csv(bind_rows(existing_log, sr_log_row), log_path)
}
```

## 8.2 SR 2023-10-09 (upstream): same treatment, second station

This logger runs continuously for daily metabolism monitoring (Section
6 note above), so its raw file spans a full ~8 h deployment rather than
a tight slug window; the whole thing is kept, since `drifting_background()`
and the 5%-of-peak return cutoff in `02` handle a long flat pre/post
baseline without distorting the integral, and trimming it by hand here
would just be guessing at a window `03` review can check visually
instead. `reach_length_m` is overridden to 52 m at export (the 72 m
inherited via `event_metadata`, shared with the downstream row for the
reason noted in Section 6, is wrong for this station).

```{r sr-manual-trim-upstream}
sr_window_up <- hobo_with_meta %>%
  filter(stream == "SR", station == "upstream", date == as.Date("2023-10-09")) %>%
  filter(!is.na(full_range_uscm), !is.na(temp_c)) %>%
  arrange(datetime) %>%
  transmute(
    source_file, date_raw, time_raw, full_range_uscm, temp_c,
    spc_calculated_uscm = spc_uscm,
    coupler_detached = NA_real_, coupler_attached = NA_real_,
    host_connected = NA_real_, stopped = NA_real_, end_of_file = NA_real_,
    stream, station, date, logger_serial, datetime, spc_uscm,
    nacl_mass_g, amonium_added_g, phophate_added_g,
    reach_length_m = 52, reach_mean_width_m,
    addition_time, addition_datetime,
    background_spc = background_spc_probe
  )

nrow(sr_window_up); range(sr_window_up$datetime)

write_csv(sr_window_up, here("data", "slug_trimmed", "slug_SR_20231009_upstream.csv"))

sr_log_row_up <- tibble(
  source_file   = "SpC_SR_091023_upstream.xlsx",
  stream        = "SR",
  date          = as.Date("2023-10-09"),
  station       = "upstream",
  start_datetime = min(sr_window_up$datetime),
  end_datetime   = max(sr_window_up$datetime),
  n_obs          = nrow(sr_window_up),
  output_file    = "slug_SR_20231009_upstream.csv",
  label          = "built directly from the raw export (Shiny app not run this session); see Sec. 8.2"
)

existing_log <- read_csv(log_path, show_col_types = FALSE)
if (!sr_log_row_up$output_file %in% existing_log$output_file) {
  write_csv(bind_rows(existing_log, sr_log_row_up), log_path)
}
```

```{r selection-log}
read_csv(here("data", "slug_trimmed_log.csv"), show_col_types = FALSE)
```

# 9. Import trimmed windows and anchor elapsed time

```{r import-trimmed}
trimmed_files <- list.files(here("data", "slug_trimmed"),
                            pattern = "\\.csv$", full.names = TRUE)

hobo_slug <- trimmed_files %>%
  set_names(basename(.)) %>%
  map(~ read_csv(.x, show_col_types = FALSE)) %>%
  bind_rows(.id = "trimmed_file") %>%
  rename(any_of(c(background_spc_probe = "background_spc"))) %>%
  mutate(
    datetime   = with_tz(datetime, tzone = "America/Sao_Paulo"),
    slug_label = str_extract(trimmed_file, "(?<=_)[PN](?=\\.csv$)")
  )

names(hobo_slug)

hobo_slug %>% count(trimmed_file, stream, date, station, slug_label)
```

> **QAQC note.** `write_csv()`/`read_csv()` preserve the correct absolute
> instant when round-tripping `POSIXct`, but `read_csv()` re-labels the
> timezone as UTC on import. Because Brasília is UTC−3, displayed times
> appeared three hours late even though elapsed-time arithmetic remained
> correct. `with_tz()` restores the display timezone without shifting the
> underlying instant.

```{r reanchor}
hobo_slug <- hobo_slug %>%
  # drop the event-level anchor inherited from the trimmed export and
  # replace it with the slug-level anchor from Section 7
  select(-any_of(c("addition_datetime", "addition_time"))) %>%
  left_join(addition_anchor, by = c("stream", "date", "slug_label")) %>%
  mutate(
    time_since_release_s = as.numeric(difftime(datetime, addition_datetime,
                                               units = "secs")),
    event_id = paste(
      stream, format(date, "%Y%m%d"),
      coalesce(slug_label, station, "single"), sep = "_"
    )
  )

# must be empty: every trimmed window resolves to exactly one anchor
hobo_slug %>% filter(is.na(addition_datetime)) %>% distinct(trimmed_file)
```

```{r check-event-id}
hobo_slug %>%
  distinct(trimmed_file, event_id) %>%
  mutate(matches_filename = event_id == str_remove(str_remove(trimmed_file, "^slug_"), "\\.csv$"))
```

# 10. QAQC checks on the trimmed series

```{r qaqc-window-summary}
window_summary <- hobo_slug %>%
  group_by(event_id, stream, date, station, slug_label) %>%
  summarise(
    n_obs            = n(),
    start            = min(datetime),
    end              = max(datetime),
    duration_min     = as.numeric(difftime(max(datetime), min(datetime), units = "mins")),
    t_start_s        = min(time_since_release_s),
    t_end_s          = max(time_since_release_s),
    dt_median_s      = median(as.numeric(diff(sort(datetime))), na.rm = TRUE),
    dt_max_s         = max(as.numeric(diff(sort(datetime))), na.rm = TRUE),
    spc_min          = min(spc_uscm, na.rm = TRUE),
    spc_max          = max(spc_uscm, na.rm = TRUE),
    spc_at_end       = last(spc_uscm[order(datetime)]),
    t_peak_s         = time_since_release_s[which.max(spc_uscm)],
    n_na_spc         = sum(is.na(spc_uscm)),
    n_before_release = sum(time_since_release_s < 0),
    .groups = "drop"
  )

window_summary
```

```{r qaqc-flags}
window_flags <- window_summary %>%
  transmute(
    event_id,
    # window opens after the release: no pre-slug baseline captured
    flag_no_pre_release = n_before_release == 0,
    # peak sits in the first or last 10 % of the window: likely clipped
    flag_peak_at_edge   = t_peak_s < t_start_s + 0.10 * (t_end_s - t_start_s) |
                          t_peak_s > t_start_s + 0.90 * (t_end_s - t_start_s),
    # series still well above its own minimum at the end: truncated tail,
    # which biases the integral used for discharge and residence time
    flag_truncated_tail = (spc_at_end - spc_min) > 0.15 * (spc_max - spc_min),
    # logging gap far above the nominal interval
    flag_logging_gap    = dt_max_s > 5 * dt_median_s,
    flag_na_present     = n_na_spc > 0
  )

window_flags

window_flags %>%
  pivot_longer(-event_id, names_to = "flag", values_to = "raised") %>%
  filter(raised) %>%
  arrange(event_id)
```

> **QAQC note.** `flag_truncated_tail` matters most for this dataset:
> transient-storage inverse modelling is disproportionately sensitive to
> the falling limb, and a window closed before the tracer returns to
> background will bias the storage parameters. Flagged events are not
> dropped here — they are carried into `02_integration.Rmd` as an
> explicit column so the decision to extrapolate, truncate, or exclude
> is made once, in the open, at the analysis stage.

```{r plot-trimmed-review}
p_trim <- hobo_slug %>%
  ggplot(aes(x = time_since_release_s / 60, y = spc_uscm)) +
  geom_line(linewidth = 0.3) +
  geom_hline(aes(yintercept = background_spc_probe),
             linetype = "dashed", colour = "firebrick") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "darkgreen") +
  facet_wrap(~ event_id, scales = "free") +
  labs(x = "Time since release (min)", y = "SpC (µS/cm)",
       title = "Trimmed slug windows",
       subtitle = "Green = release | Red = handheld probe background") +
  theme_minimal(base_size = 10)

ggplotly(p_trim)
```

# 11. Export

```{r export-btc}
dir.create(here("data_derived", "dictionary"), recursive = TRUE, showWarnings = FALSE)

btc_conservative <- hobo_slug %>%
  left_join(window_flags, by = "event_id") %>%
  transmute(
    event_id, stream, date, station, slug_label,
    datetime, time_since_release_s,
    full_range_uscm, temp_c, spc_uscm,
    background_spc_probe,
    logger_serial, source_file, trimmed_file,
    flag_no_pre_release, flag_peak_at_edge,
    flag_truncated_tail, flag_logging_gap, flag_na_present
  ) %>%
  arrange(event_id, datetime)

events_hobo <- hobo_slug %>%
  distinct(event_id, stream, date, station, slug_label,
           addition_datetime, label,
           nacl_mass_g, amonium_added_g, phophate_added_g,
           reach_length_m, reach_mean_width_m,
           background_spc_probe, logger_serial, source_file) %>%
  left_join(window_summary %>% select(event_id, n_obs, start, end, duration_min,
                                      dt_median_s, spc_min, spc_max, t_peak_s),
            by = "event_id") %>%
  left_join(window_flags, by = "event_id") %>%
  arrange(stream, date)

write_csv(btc_conservative, here("data_derived", "btc_conservative.csv"))
write_csv(events_hobo,      here("data_derived", "events_hobo.csv"))
saveRDS(btc_conservative,   here("data_derived", "btc_conservative.rds"))
saveRDS(events_hobo,        here("data_derived", "events_hobo.rds"))

dim(btc_conservative); dim(events_hobo)
```

```{r export-dictionary}
dict_btc <- tribble(
  ~column, ~unit, ~description,
  "event_id",             "-",     "Unique event key: stream_YYYYMMDD_{slug_label|station|single}. Primary key linking all tables in this project.",
  "stream",               "-",     "Stream code",
  "date",                 "-",     "Event date (Brasília time)",
  "station",              "-",     "upstream / downstream for two-station slugs, NA otherwise",
  "slug_label",           "-",     "P / N for the split RA 2023-09-06 release, NA otherwise",
  "datetime",             "-",     "Logger timestamp, America/Sao_Paulo (UTC-3)",
  "time_since_release_s", "s",     "Elapsed time from the tracer addition anchoring this slug",
  "full_range_uscm",      "uS/cm", "Uncompensated full-range conductivity as logged",
  "temp_c",               "degC",  "Water temperature",
  "spc_uscm",             "uS/cm", "Specific conductance, recalculated as full_range / (1 + 0.02*(temp_c - 25))",
  "background_spc_probe", "uS/cm", "Pre-release SpC from the handheld probe; independent reference, not a logger value",
  "logger_serial",        "-",     "HOBO logger serial number",
  "source_file",          "-",     "Raw .xlsx export the observation came from",
  "trimmed_file",         "-",     "Trimmed CSV written by the Shiny selection app",
  "flag_no_pre_release",  "-",     "TRUE if the window contains no observation before the release",
  "flag_peak_at_edge",    "-",     "TRUE if the SpC peak falls in the outer 10% of the window",
  "flag_truncated_tail",  "-",     "TRUE if SpC at the end of the window is still >15% above the window minimum",
  "flag_logging_gap",     "-",     "TRUE if the largest inter-record gap exceeds 5x the median logging interval",
  "flag_na_present",      "-",     "TRUE if any SpC value is missing"
)

write_csv(dict_btc, here("data_derived", "dictionary", "dict_btc_conservative.csv"))
dict_btc
```

```{r session-info}
sessionInfo()
```
