---
title: "02 — Integration and hydraulic characterisation"
subtitle: "Nutrient grab samples, conservative tracer, and event-level hydraulics"
author: "Kauan Fonseca"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6
)

library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
library(lubridate)
library(ggplot2)
library(plotly)
library(pracma)
library(here)
```

```{r constants}
# mg/L NaCl per uS/cm of background-corrected SpC; laboratory calibration.
NACL_SLOPE <- 0.461159
```

# Scope

Takes the instrument-clean logger series from `01_hobo_qaqc.Rmd` and the
revised nutrient grab samples, and builds the analysis-ready tables that
feed the inverse transient-storage and nutrient-uptake modelling.

Covers: timestamp reconstruction of the grab samples, harmonisation to
long format, a time-varying baseline for NaCl, pairing of each grab to
the interpolated logger series, probe-vs-logger reconciliation,
event-level hydraulics, and the CaCO₃ co-precipitation correction for
SRP.

Hysteresis and the final BTC review live in `03_btc_review.Rmd`.

**Outputs:** `data_derived/btc_nutrients.csv`,
`data_derived/btc_conservative.csv`, `data_derived/events.csv`,
`data_derived/geochem_check.csv`.

```{r read-01}
btc_conservative <- read_rds(here("data_derived", "btc_conservative.rds"))
events_hobo      <- read_rds(here("data_derived", "events_hobo.rds"))

# RA 2023-09-06 N slug used 6240 g NaCl, not the 5957 g inherited from the
# combined field-sheet row.
events_hobo <- events_hobo %>%
  mutate(nacl_mass_g = if_else(event_id == "RA_20230906_N", 6240, nacl_mass_g))

btc_conservative %>% distinct(event_id, stream, date, station, slug_label)
```

```{r flag-drop}
btc_conservative <- btc_conservative %>%
  group_by(event_id) %>%
  mutate(
    baseline_ref = {
      pre <- spc_uscm[time_since_release_s < 0 &
                      spc_uscm > 0.5 * median(spc_uscm, na.rm = TRUE)]
      if (length(pre) >= 3) median(pre, na.rm = TRUE)
      else median(spc_uscm, na.rm = TRUE)
    },
    out_of_water = !is.na(baseline_ref) & spc_uscm < 0.5 * baseline_ref
  ) %>%
  ungroup()

btc_conservative %>%
  filter(out_of_water) %>%
  group_by(event_id) %>%
  summarise(n_flagged = n(), spc_min = min(spc_uscm),
            all_pre = all(time_since_release_s < 0), .groups = "drop")

btc_conservative <- btc_conservative %>%
  filter(!out_of_water) %>%
  select(-baseline_ref, -out_of_water)

btc_conservative %>% distinct(event_id) %>% nrow()   # esperado: 13 (11 + SR_20231009_downstream + SR_20231009_upstream)
```

> **QAQC note.** Events whose window opens after the release have no
> pre-release stretch, so the baseline reference falls back to the event
> median. Without that fallback the flag is `NA` and the whole event is
> dropped by the filter. The row count is the guard.

------------------------------------------------------------------------

# 1. Import nutrient grab samples

```{r read-nutrients}
read_nutrient <- function(path) {
  read_csv(path, show_col_types = FALSE) %>%
    mutate(
      date = ymd(as.character(date)),
      datetime_field = with_tz(
        as.POSIXct(paste(date, time_along_btc), tz = "America/Cuiaba"),
        "America/Sao_Paulo"),
      release_datetime_field = with_tz(
        as.POSIXct(paste(date, release_time), tz = "America/Cuiaba"),
        "America/Sao_Paulo")
    )
}

n_raw <- read_nutrient(here("data", "nutrients", "nutrient_addition_nitrogen_data.csv"))
p_raw <- read_nutrient(here("data", "nutrients", "nutrient_addition_phosphate_data.csv"))

dim(n_raw); dim(p_raw)
```

```{r diagnose-release-times}
bind_rows(n_raw %>% mutate(solute = "NH4-N"),
          p_raw %>% mutate(solute = "SRP")) %>%
  mutate(diff_s = as.numeric(difftime(datetime_field, release_datetime_field,
                                      units = "secs")) - time_since_addition_s) %>%
  count(stream, date, solute, diff_s) %>%
  arrange(desc(abs(diff_s)))
```

> **QAQC note.** Three events fail this check. RA 2023-09-06 (SRP) is
> off by −8100 s because P and N were released sequentially (P at 14:30,
> N at 16:45) and the phosphate sheet carries the nitrogen release time;
> the phosphate curve is anchored to 14:30. CB 2023-09-07 (+180 s) and
> RS/SR 2023-10-11 (+40 s / +10 s) reflect rounding in the field
> notebook. `time_since_addition_s` is the reliable field, so timestamps
> are reconstructed as `addition_datetime + time_since_addition_s` and
> the discrepancy kept in `anchor_drift_s`. This compares two ways of
> dating a grab within the spreadsheet only; it cannot detect an offset
> between the field clock and the logger clock.

# 2. Harmonise to long format

N and P at CB 2023-09-07 were added together but sampled alternately, so
they are distinct samples at distinct times. The two tables are stacked,
not joined.

```{r harmonise}
common_cols <- function(df, solute_name, conc, background, conc_corr) {
  df %>%
    transmute(stream, date, solute = solute_name,
              datetime_field, release_datetime_field, time_since_addition_s,
              sample_btc, curve_limb,
              conc_ugL = {{ conc }}, background_ugL = {{ background }},
              conc_corr_ugL = {{ conc_corr }},
              spc_uscm_grab      = SpC_uScm,
              spc_corr_uscm_grab = corr_SpC_uScm,
              nacl_mgL_grab_src  = NaCl_mgL)
}

nutrients_long <- bind_rows(
  common_cols(n_raw, "NH4-N", measured_NH4_N_ug_L, nh4_n_ugL_bck, bck_corrected_N_ug_L),
  common_cols(p_raw, "SRP",   measured_SRP_ug_L,   srp_ugL_bck,   bck_corrected_SRP_ug_L)
) %>%
  mutate(nacl_mgL_grab = NACL_SLOPE * pmax(spc_corr_uscm_grab, 0)) %>%
  arrange(stream, date, solute, time_since_addition_s)

nutrients_long %>% count(stream, date, solute)
```

> **QAQC note.** `nacl_mgL_grab` is recomputed from the spreadsheet's
> background-corrected SpC so the same conversion applies to both
> instruments; the original column is kept as `nacl_mgL_grab_src`. Both
> rest on a constant background, unlike the logger series of Section 3.

```{r assign-event-id}
nutrients_long <- nutrients_long %>%
  mutate(
    slug_label = case_when(
      stream == "RA" & date == as.Date("2023-09-06") & solute == "SRP"   ~ "P",
      stream == "RA" & date == as.Date("2023-09-06") & solute == "NH4-N" ~ "N",
      TRUE ~ NA_character_),
    station = if_else(stream == "RA" & date == as.Date("2023-10-05"),
                      "downstream", NA_character_),
    event_id = paste(stream, format(date, "%Y%m%d"),
                     coalesce(slug_label, station, "single"), sep = "_")
  ) %>%
  left_join(events_hobo %>% select(event_id, addition_datetime), by = "event_id") %>%
  mutate(
    datetime = coalesce(addition_datetime + time_since_addition_s, datetime_field),
    anchor_drift_s = as.numeric(difftime(datetime_field, datetime, units = "secs"))
  )

nutrients_long %>%
  group_by(event_id, solute) %>%
  summarise(n = n(), t0 = min(datetime), .groups = "drop")
```

```{r key-reconciliation}
nutrients_long %>%
  distinct(event_id) %>% mutate(has_nutrients = TRUE) %>%
  full_join(btc_conservative %>% distinct(event_id) %>% mutate(has_logger = TRUE),
            by = "event_id") %>%
  arrange(event_id)
```

> **QAQC note.** `SR_20231011_single` has grabs but no logger series:
> usable for nutrients and for a handheld discharge estimate, not for
> inverse modelling. `RA_20231005_upstream` has a logger series but no
> grabs, which is correct — sampling was downstream only.

# 3. NaCl from specific conductance, with a time-varying baseline

Background specific conductance drifts through the experiment with
temperature and photosynthetic CO₂ drawdown, so a single constant
baseline drives the tail below zero and corrupts the integral. The
baseline is modelled as a straight line between the pre-release
conductance and the conductance at the end of the window, subtracted
point by point.

The end anchor is valid because the logger was retrieved once SpC had
returned to its pre-release value. The exception is the events flagged
`flag_truncated_tail` in `01`, where the logger was pulled before full
return. For those the end anchor is the lowest point **after the peak**,
anchored at its own time. Taking the minimum of the whole window and
pinning it to the end of the window tilts the baseline across the entire
tail whenever that minimum falls on the rising limb, which is what drove
CB 2023-10-08 to an implausibly low discharge in earlier versions;
restricting the search to strictly after the peak is what fixes that.

An earlier version of this fix additionally capped the end anchor so it
could never sit above the start anchor — treating any post-peak minimum
higher than the pre-release baseline as suspect. That is too strong: CB
2023-10-08's background genuinely drifts upward over the event (the
smoothed post-peak anchor comes out at \~410.7 µS/cm against a
pre-release baseline of \~409.0, a drift confirmed by the raw record —
the tail sits on a stable \~410-411 µS/cm plateau, not at 409), and the
cap discarded that real drift, flattening the correction and inflating
the tail integral. Removing the cap and letting the post-peak anchor
stand restores CB 2023-10-08 to \~315 L/s (against the \~317 on record)
with no change to CD 2023-10-09 (its background never triggered the cap;
it uses `PRE_STABLE`, not this truncated-tail branch at all) or to CB
2023-09-07 (its post-peak anchor already sits below the pre-release
baseline).

The tradeoff is explicit: `RA_20231005_downstream` shares CB
2023-10-09's pre-tracer instrument step (its start anchor drifts
continuously from 486 to the low 470s µS/cm over the \~700 s before the
plateau, unlike CD's single-interval step) and has no `PRE_STABLE`
window of its own yet (§5.6, not blocking — the event uses the manual
discharge value). Without the cap its post-peak anchor, computed against
that unreliable start anchor, produces a nonsensical
`discharge_Ls_logger` in the diagnostic table. That column was already
flagged unreliable for this event and is not the adopted value, so this
is a known caveat made more visible, not a new problem.

```{r drifting-background-fn}
# Events where the pre-release window does not represent the baseline during
# tracer passage. CD 2023-10-09 steps down 12 uS/cm in a single 10 s interval
# at constant temperature, ~160 s after release and well before the tracer
# arrives: an instrument transition, not water.
PRE_STABLE <- list(
  "CD_20231009_single" = c(300, 780)
  # "RA_20231005_downstream" = c(?, ?)   # same kind of step; window not identified
)

drifting_background <- function(df, n_anchor = 10, pre_stable_s = NULL) {
  df <- df %>% arrange(time_since_release_s)
  truncated <- isTRUE(df$flag_truncated_tail[1])

  pre <- if (!is.null(pre_stable_s)) {
    df %>% filter(time_since_release_s >= pre_stable_s[1],
                  time_since_release_s <= pre_stable_s[2])
  } else {
    df %>% filter(time_since_release_s < 0)
  }

  if (nrow(pre) >= 3) {
    y0 <- median(pre$spc_uscm, na.rm = TRUE)
    t0 <- max(pre$time_since_release_s)
  } else {
    y0 <- median(head(df$spc_uscm, n_anchor), na.rm = TRUE)
    t0 <- min(df$time_since_release_s)
  }

  if (truncated) {
    peak_i <- which.max(df$spc_uscm)
    post   <- seq_along(df$spc_uscm) > peak_i
    j <- if (!any(post)) which.min(df$spc_uscm) else which(post)[which.min(df$spc_uscm[post])]
    k  <- max(1, j - n_anchor %/% 2):min(nrow(df), j + n_anchor %/% 2)
    y1 <- median(df$spc_uscm[k], na.rm = TRUE)
    t1 <- median(df$time_since_release_s[k], na.rm = TRUE)
    # No cap against y0 here (removed): a post-peak anchor above the
    # pre-release baseline can be a real upward drift (CB 2023-10-08), not
    # only a mis-identified minimum. The peak-restricted search above is
    # what guards against the original bug; see the prose note above this
    # chunk for the CB 2023-10-08 / RA_20231005_downstream tradeoff.
    anchor_end <- "post_peak_min"
  } else {
    tail_df <- tail(df, n_anchor)
    y1 <- median(tail_df$spc_uscm, na.rm = TRUE)
    t1 <- median(tail_df$time_since_release_s, na.rm = TRUE)
    anchor_end <- "tail_return"
  }

  df$background_drift_uscm <- approx(c(t0, t1), c(y0, y1),
                                     xout = df$time_since_release_s, rule = 2)$y
  df$background_anchor_end <- anchor_end
  df$background_start_src  <- if (is.null(pre_stable_s)) "pre_release" else "stable_window"
  df
}

btc_conservative <- btc_conservative %>%
  select(-any_of(c("background_drift_uscm", "background_anchor_end",
                   "background_start_src", "spc_corr_uscm", "nacl_mgL"))) %>%
  group_by(event_id) %>%
  group_modify(~ drifting_background(.x, pre_stable_s = PRE_STABLE[[.y$event_id]])) %>%
  ungroup() %>%
  mutate(
    spc_corr_uscm = spc_uscm - background_drift_uscm,
    nacl_mgL      = NACL_SLOPE * pmax(spc_corr_uscm, 0)
  )

btc_conservative %>% distinct(event_id) %>% nrow()              # esperado: 13 (11 + SR_20231009_downstream + SR_20231009_upstream)
btc_conservative %>% summarise(min_nacl = min(nacl_mgL, na.rm = TRUE))
btc_conservative %>% distinct(event_id, background_anchor_end, background_start_src)
```

```{r drifting-background-check}
btc_conservative %>%
  filter(time_since_release_s >= 0) %>%
  group_by(event_id) %>%
  summarise(
    anchor_end = first(background_anchor_end),
    frac_neg   = mean(spc_corr_uscm < -0.5),
    frac_zero  = mean(nacl_mgL == 0),
    drift_uscm = last(background_drift_uscm) - first(background_drift_uscm),
    integral_nacl = pracma::trapz(time_since_release_s, pmax(spc_corr_uscm, 0)) * NACL_SLOPE,
    .groups = "drop"
  ) %>%
  arrange(desc(frac_neg))
```

> **QAQC note.** `frac_zero` is the diagnostic that matters for weak
> events: where the peak is only a few percent above baseline, natural
> variability is of the same order as the signal in the tail, so
> flooring removes real curve area and inflates discharge. No baseline
> fixes this, so those events are handled in Section 6 by declaring a
> different discharge source rather than by tuning the background.

# 4. Pair each grab to the logger series

```{r pair-function}
pair_to_logger <- function(grabs, logger, max_gap_s = 60) {
  grabs %>%
    group_by(event_id) %>%
    group_modify(function(g, key) {
      l <- logger %>% filter(event_id == key$event_id) %>% arrange(datetime)
      if (nrow(l) < 2)
        return(mutate(g, spc_uscm_logger = NA_real_,
                      nacl_mgL_logger = NA_real_, gap_s = NA_real_))
      tl <- as.numeric(l$datetime); tg <- as.numeric(g$datetime)
      g$spc_uscm_logger <- approx(tl, l$spc_uscm, xout = tg, rule = 1, ties = mean)$y
      g$nacl_mgL_logger <- approx(tl, l$nacl_mgL, xout = tg, rule = 1, ties = mean)$y
      g$gap_s <- vapply(tg, function(t) min(abs(tl - t)), numeric(1))
      g
    }) %>%
    ungroup() %>%
    mutate(across(c(spc_uscm_logger, nacl_mgL_logger),
                  ~ if_else(is.na(gap_s) | gap_s > max_gap_s, NA_real_, .x)))
}

paired <- pair_to_logger(nutrients_long, btc_conservative)

paired %>%
  group_by(event_id, solute) %>%
  summarise(n = n(), n_matched = sum(!is.na(nacl_mgL_logger)),
            gap_median_s = median(gap_s, na.rm = TRUE), .groups = "drop")
```

```{r nacl-source}
paired <- paired %>%
  mutate(
    nacl_mgL = coalesce(nacl_mgL_logger, nacl_mgL_grab),
    nacl_source = case_when(
      !is.na(nacl_mgL_logger) ~ "logger_interpolated",
      !is.na(nacl_mgL_grab)   ~ "handheld_grab",
      TRUE                    ~ NA_character_)
  )

paired %>% count(event_id, nacl_source)
```

> **QAQC note.** Merging is an added column, not a second table. Both
> `nacl_mgL_grab` and `nacl_mgL_logger` are kept; `nacl_mgL` carries the
> value used downstream and `nacl_source` records which one. `gap_s` is
> interpolation quality, not alignment.

# 5. Probe vs logger reconciliation

```{r offset-regression}
paired %>%
  filter(!is.na(spc_uscm_logger)) %>%
  group_by(event_id) %>% filter(n() >= 4) %>%
  summarise(n = n(),
            slope = coef(lm(spc_uscm_grab ~ spc_uscm_logger))[2],
            r2 = summary(lm(spc_uscm_grab ~ spc_uscm_logger))$r.squared,
            mean_diff = mean(spc_uscm_grab - spc_uscm_logger),
            .groups = "drop") %>%
  arrange(r2)
```

```{r residual-by-limb}
resid_limb <- paired %>%
  filter(!is.na(spc_uscm_logger)) %>%
  mutate(resid = spc_uscm_grab - spc_uscm_logger,
         curve_limb = factor(curve_limb, levels = c("rising", "peak", "falling")))

ggplotly(
  ggplot(resid_limb, aes(curve_limb, resid, fill = curve_limb)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(alpha = 0.6, outlier.size = 0.6) +
    facet_wrap(~ event_id, scales = "free_y") +
    labs(x = NULL, y = "Probe − logger (µS/cm)") +
    theme_minimal(base_size = 9) + theme(legend.position = "none")
)
```

> **QAQC note.** The two instruments sat close together at the same
> station, so no offset correction is warranted. Residuals separate by
> limb because the tracer cloud is hysteretic: rising and falling limbs
> carry different concentrations at a given instant, and two sensors a
> short distance apart sample slightly different points of that field.
> Tracer dynamics, not a calibration fault. Left uncorrected.

```{r peak-ratio}
btc_conservative %>%
  group_by(event_id) %>%
  summarise(baseline = median(spc_uscm[time_since_release_s < 0], na.rm = TRUE),
            spc_max = max(spc_uscm, na.rm = TRUE),
            peak_ratio = spc_max / baseline,
            .groups = "drop") %>%
  left_join(events_hobo %>%
              transmute(event_id, mixing_lengths = reach_length_m / reach_mean_width_m),
            by = "event_id") %>%
  arrange(desc(peak_ratio))
```

> **QAQC note.** Karst events sit at a peak ratio of 1.03–1.29.
> `RA_20231005_upstream` reaches 3.61: that logger sat close to the
> release point in a bedrock riffle and the salt had not dispersed
> across the section, so dilution gauging's premise fails there.
> `SA_20231010` is high for a benign reason — its baseline is \~4.9
> µS/cm, so any tracer gives a large ratio. CD 2023-10-09 has only \~5.7
> mixing lengths against the \~20 usually recommended.

# 6. Event-level hydraulics

Discharge by integral (slug) dilution gauging:

$$ Q = \frac{M}{\displaystyle\int_0^{\infty} C(t)\,dt} $$

For the **logger** the integral is trapezoidal over the high-resolution
record with the drifting baseline of Section 3, and $t_p$ is the time to
the true SpC peak. For the **handheld** series the integral is
trapezoidal over the sparse grabs and $t_p$ the end of the rising limb,
reproducing the original manual workflow. Velocity is $v = L / t_p$;
depth follows from continuity, $z = Q / (v\,w)$.

The logger is the default source. Where its integral is invalid for a
documented physical reason, the value is taken from the handheld series
or the manual calculation, and the source is recorded per event.

```{r reach-geometry}
# Authoritative reach geometry, resolved against field survey records.
reach_geometry <- tribble(
  ~stream, ~date,                 ~reach_length_m, ~reach_mean_width_m,
  "CB",    as.Date("2023-09-07"), 162,             7.75,
  "CB",    as.Date("2023-10-08"), 162,             7.75,
  "RA",    as.Date("2023-09-06"), 135,             9.46,
  "RA",    as.Date("2023-10-05"), 135,             9.46,
  "CD",    as.Date("2023-10-09"),  52,             9.20
)

# Events with grabs but no logger never enter events_hobo, which is built from
# the trimmed logger CSVs, so their mass and geometry come from the field sheet.
# Without this they carry NA mass and no discharge can be computed at all.
GEOMETRY_NO_LOGGER <- tribble(
  ~event_id,             ~nacl_mass_g, ~reach_length_m, ~reach_mean_width_m,
  "SR_20231011_single",  638.9,        72,              1.5
)

geometry <- events_hobo %>%
  distinct(event_id, stream, date, nacl_mass_g,
           reach_length_m, reach_mean_width_m) %>%
  left_join(reach_geometry, by = c("stream", "date"), suffix = c("", "_auth")) %>%
  mutate(reach_length_m     = coalesce(reach_length_m_auth, reach_length_m),
         reach_mean_width_m = coalesce(reach_mean_width_m_auth, reach_mean_width_m)) %>%
  select(event_id, nacl_mass_g, reach_length_m, reach_mean_width_m) %>%
  bind_rows(GEOMETRY_NO_LOGGER) %>%
  distinct(event_id, .keep_all = TRUE)

# guard: nenhum evento com grabs pode ficar sem massa ou geometria
geometry %>%
  filter(event_id %in% nutrients_long$event_id) %>%
  filter(if_any(c(nacl_mass_g, reach_length_m, reach_mean_width_m), is.na))
```

```{r hydraulics-logger}
integrate_to_return <- function(t, nacl, spc, return_frac = 0.05) {
  o <- order(t); t <- t[o]; nacl <- nacl[o]; spc <- spc[o]
  peak_i <- which.max(spc)
  peak_nacl <- nacl[peak_i]
  after <- seq_along(nacl) > peak_i
  below <- after & nacl < return_frac * peak_nacl
  end_i <- if (any(below)) which(below)[1] else length(nacl)
  list(
    integral = pracma::trapz(t[1:end_i], pmax(nacl[1:end_i], 0)),
    t_end_s  = t[end_i],
    t_peak_s = t[peak_i]
  )
}

hydro_logger <- btc_conservative %>%
  filter(time_since_release_s >= 0, !is.na(nacl_mgL)) %>%
  group_by(event_id) %>%
  group_modify(function(d, key) {
    r <- integrate_to_return(d$time_since_release_s, d$nacl_mgL, d$spc_uscm)
    tibble(integral_Ct_mgLs = r$integral, t_peak_s = r$t_peak_s,
           t_return_s = r$t_end_s, truncated_tail = first(d$flag_truncated_tail))
  }) %>%
  ungroup() %>%
  left_join(geometry, by = "event_id") %>%
  mutate(
    discharge_Ls_logger      = (nacl_mass_g * 1000) / integral_Ct_mgLs,
    water_velocity_ms_logger = reach_length_m / t_peak_s,
    stream_depth_m_logger    = (discharge_Ls_logger / 1000) /
                               (water_velocity_ms_logger * reach_mean_width_m)
  )

hydro_logger %>%
  select(event_id, integral_Ct_mgLs, t_peak_s, t_return_s,
         discharge_Ls_logger, truncated_tail)
```

```{r hydraulics-probe}
# Sampling gaps. The handheld integral is trapezoidal, so a long unsampled
# stretch is filled by a straight line between the bracketing grabs. Where one
# interval carries a large share of the total area the estimate depends more on
# that interpolation than on the data, and the flag says so.
gap_diag <- nutrients_long %>%
  filter(!is.na(nacl_mgL_grab)) %>%
  distinct(event_id, time_since_addition_s, nacl_mgL_grab) %>%
  arrange(event_id, time_since_addition_s) %>%
  group_by(event_id) %>%
  mutate(dt = lead(time_since_addition_s) - time_since_addition_s,
         area_trap = (nacl_mgL_grab + lead(nacl_mgL_grab)) / 2 * dt) %>%
  summarise(max_gap_s = max(dt, na.rm = TRUE),
            frac_area_max_gap = max(area_trap, na.rm = TRUE) /
                                sum(area_trap, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(flag_sparse_recession = frac_area_max_gap > 0.3)

hydro_probe <- nutrients_long %>%
  filter(!is.na(nacl_mgL_grab)) %>%
  distinct(event_id, time_since_addition_s, curve_limb, nacl_mgL_grab) %>%
  arrange(event_id, time_since_addition_s) %>%
  group_by(event_id) %>%
  summarise(
    integral_Ct_mgLs = if (n() >= 3 && diff(range(time_since_addition_s)) > 0)
                         pracma::trapz(time_since_addition_s, pmax(nacl_mgL_grab, 0))
                       else NA_real_,
    t_peak_s = if (any(curve_limb == "rising"))
                 max(time_since_addition_s[curve_limb == "rising"]) else NA_real_,
    .groups = "drop"
  ) %>%
  left_join(geometry, by = "event_id") %>%
  left_join(gap_diag, by = "event_id") %>%
  mutate(
    discharge_Ls_probe      = if_else(is.na(integral_Ct_mgLs) | integral_Ct_mgLs <= 0,
                                      NA_real_, (nacl_mass_g * 1000) / integral_Ct_mgLs),
    water_velocity_ms_probe = reach_length_m / t_peak_s,
    stream_depth_m_probe    = (discharge_Ls_probe / 1000) /
                              (water_velocity_ms_probe * reach_mean_width_m)
  )

nutrients_long %>%
  group_by(event_id) %>%
  summarise(n_nacl   = sum(!is.na(nacl_mgL_grab)),
            n_rising = sum(curve_limb == "rising", na.rm = TRUE),
            t_span_s = diff(range(time_since_addition_s, na.rm = TRUE)),
            .groups = "drop") %>%
  arrange(n_nacl)

hydro_probe %>%
  select(event_id, integral_Ct_mgLs, t_peak_s, discharge_Ls_probe,
         max_gap_s, frac_area_max_gap, flag_sparse_recession)
```

> **QAQC note.** `SR_20231011_single` carries the only serious gap: the
> recession was left unsampled for 1350 s starting 1230 s after the
> peak, and that single interval holds a large share of the trapezoidal
> area. Filling it with a left-rectangle rule instead — holding the last
> concentration flat until the next grab — gives 2.9 L/s rather than
> 4.4, but a flat plateau followed by an abrupt drop is not how a tracer
> cloud passes, so the trapezoid is retained. The value is supported
> independently by the same reach two days earlier, where a logger that
> ran to full baseline return gave \~4.3 L/s from a different tracer
> mass.

```{r hydraulics-combine}
# Discharge source per event. Default is the logger; this table records only
# the events whose logger integral is invalid, with the reason.
#   probe  -> handheld trapezoidal estimate
#   manual -> manual Excel value (fixed background, negatives retained);
#             no event currently uses this as its adopted source -- kept as
#             a reference value (discharge_Ls_manual) for every event it was
#             ever entered for, and as a live fallback if a future logger AND
#             probe estimate both fail for some event.
#   none   -> no valid estimate; discharge stays NA
#
# RA_20230906_N/_P and RA_20231005_downstream were manual until Kauan's
# correction: each has a documented physical concern (no pre-release
# baseline, a peak barely above background, a contaminated start anchor)
# that argued against trusting an automated trapezoidal estimate there --
# but that concern is now carried forward as an explicit flag on the
# adopted probe value (flag_no_baseline, flag_weak_signal below), not as a
# reason to fall back to a hand-picked number instead. Whether to trust,
# down-weight, or discard these three in a downstream analysis is a
# modelling decision made against the flag, not something decided here.
DISCHARGE_SOURCE_OVERRIDE <- tribble(
  ~event_id,                ~use_source, ~reason_discharge,
  "RA_20231005_upstream",   "none",      "incomplete mixing (peak ratio 3.61): dilution gauging premise fails",
  "RA_20231005_downstream", "probe",     "start anchor contaminated by the pre-tracer step; logger depth physically impossible",
  "RA_20230906_N",          "probe",     "no pre-release baseline; residual salt from the preceding P slug",
  "RA_20230906_P",          "probe",     "peak only 3% above baseline; background noise of the same order as the signal",
  "SR_20231011_single",     "probe",     "no logger series; recession unsampled for 1350 s from 1230 s after the peak"
)

DISCHARGE_MANUAL <- tribble(
  ~event_id,                ~discharge_Ls_manual,
  "RA_20230906_P",          251,
  "RA_20230906_N",          205,
  "RA_20231005_downstream", 201,
  "RA_20231005_upstream",   290
)

hydraulics <- hydro_logger %>%
  select(event_id, discharge_Ls_logger, water_velocity_ms_logger,
         stream_depth_m_logger, t_peak_s_logger = t_peak_s, truncated_tail) %>%
  full_join(
    hydro_probe %>%
      select(event_id, discharge_Ls_probe, water_velocity_ms_probe,
             stream_depth_m_probe, t_peak_s_probe = t_peak_s,
             max_gap_s, frac_area_max_gap, flag_sparse_recession),
    by = "event_id"
  ) %>%
  left_join(geometry,                  by = "event_id") %>%
  left_join(DISCHARGE_SOURCE_OVERRIDE, by = "event_id") %>%
  left_join(DISCHARGE_MANUAL,          by = "event_id") %>%
  mutate(
    use_source      = coalesce(use_source, "logger"),
    discharge_ratio = discharge_Ls_logger / discharge_Ls_probe,

    discharge_source = case_when(
      use_source == "none"                                 ~ NA_character_,
      use_source == "logger" & !is.na(discharge_Ls_logger) ~ "logger",
      use_source == "probe"  & !is.na(discharge_Ls_probe)  ~ "probe",
      use_source == "manual" & !is.na(discharge_Ls_manual) ~ "manual_excel",
      !is.na(discharge_Ls_probe)  ~ "probe",
      !is.na(discharge_Ls_manual) ~ "manual_excel",
      !is.na(discharge_Ls_logger) ~ "logger",
      TRUE ~ NA_character_
    ),

    discharge_Ls = case_when(
      discharge_source == "logger"       ~ discharge_Ls_logger,
      discharge_source == "probe"        ~ discharge_Ls_probe,
      discharge_source == "manual_excel" ~ discharge_Ls_manual,
      TRUE ~ NA_real_
    ),

    t_peak_s = if_else(discharge_source == "logger",
                       t_peak_s_logger,
                       coalesce(t_peak_s_probe, t_peak_s_logger)),
    water_velocity_ms = reach_length_m / t_peak_s,
    stream_depth_m    = (discharge_Ls / 1000) /
                        (water_velocity_ms * reach_mean_width_m),

    discharge_method = case_when(
      discharge_source == "logger"       ~ "logger_trapz_truepeak",
      discharge_source == "probe"        ~ "probe_trapz_endrising",
      discharge_source == "manual_excel" ~ "manual_fixed_background",
      TRUE ~ NA_character_
    ),

    flag_discharge_alt     = !is.na(discharge_source) & discharge_source != "logger",
    flag_discharge_invalid = is.na(discharge_Ls),
    flag_mixing_incomplete = event_id %in% "RA_20231005_upstream",
    flag_no_baseline       = event_id %in% c("RA_20230906_N", "RA_20231005_downstream"),
    # RA_20230906_P: probe/logger peak only ~3% above background -- the
    # trapezoidal integral is adopted (per reason_discharge above) but its
    # signal-to-noise is low, independent of flag_no_baseline (P has a
    # clean pre-release baseline; N does not).
    flag_weak_signal       = event_id %in% "RA_20230906_P",
    flag_sparse_recession  = coalesce(flag_sparse_recession, FALSE)
  ) %>%
  select(-use_source)

hydraulics %>%
  select(event_id, discharge_Ls, discharge_source, discharge_Ls_logger,
         discharge_Ls_probe, discharge_Ls_manual, discharge_ratio,
         water_velocity_ms, stream_depth_m, truncated_tail,
         flag_discharge_alt, flag_discharge_invalid, flag_no_baseline,
         flag_weak_signal, flag_sparse_recession) %>%
  arrange(event_id) %>%
  print(n = Inf)
```

```{r hydraulics-logger-vs-probe-plot}
# Every event with BOTH a logger and a handheld/probe trapezoidal estimate,
# regardless of which one ended up adopted as discharge_Ls -- so an event
# where the logger was rejected still shows how far its logger estimate sat
# from the probe one, and vice versa.
compare_lp <- hydraulics %>%
  filter(!is.na(discharge_Ls_logger), !is.na(discharge_Ls_probe))

lp_range <- range(c(compare_lp$discharge_Ls_logger, compare_lp$discharge_Ls_probe))

ggplotly(
  ggplot(compare_lp, aes(discharge_Ls_logger, discharge_Ls_probe,
                          color = discharge_source, text = event_id)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(size = 2.5) +
    scale_x_log10(limits = lp_range) +
    scale_y_log10(limits = lp_range) +
    coord_fixed() +
    labs(x = "Logger discharge (L/s)", y = "Probe discharge (L/s)",
         color = "Adopted source") +
    theme_minimal(base_size = 10),
  tooltip = "text"
)
```

> **QAQC note.** `discharge_ratio` (logger / probe) is reported for
> every event that has both, including those where the logger is not the
> adopted source, so the disagreement stays visible. Events flagged
> `truncated_tail` had the logger pulled before full baseline return, so
> their integral is a lower bound and discharge an upper bound.
> `RA_20231005_upstream` has no valid discharge: the sensor sampled the
> plume core rather than a mixed section, so neither dilution gauging
> nor transient-storage modelling is defensible there, though the curve
> remains usable for arrival time. The manual values
> (`discharge_Ls_manual`) are kept in the table as a reference only —
> they carry the bias of a fixed background with negatives retained, and
> are not the adopted source for any event as of this revision.
>
> **QAQC note — the 1:1 plot above.** Point colour is the *adopted*
> `discharge_source`, not a judgement about which axis is "right" — a
> probe-coloured point sitting near the line is not itself evidence the
> probe is correct there, only that the two independent estimates happen
> to agree. RA_20230906_N, RA_20230906_P, and RA_20231005_downstream are
> the three events where the logger was rejected (reasons in
> `reason_discharge` above) and the probe trapezoidal estimate is now
> adopted in its place; each carries a flag documenting why the logger
> reading is not trusted there (`flag_no_baseline` or
> `flag_weak_signal`) — that flag is not a case against the probe value
> itself, and no row is excluded on this basis.

## 6.1 SR 2023-10-09: two stations, and a discharge mismatch that turned

out not to be one

The `Q = 4.30 L/s` / 720,452.03 mg/L·s mass-flux integral Kauan first
gave as a field summary for "the SR slug" turned out, once traced down,
to be his own manual calculation for the **upstream** station (52 m),
not the downstream one — its point-by-point NaCl mass-flux column sums
to 720,452.0299 mg/L·s exactly. Paired against the 72 m downstream
distance in the original summary table, it looked like a direct
competing estimate for the same discharge; it is not — it is a different
station on the same reach. The comparison below is the correct one: both
stations run through the same pipeline, same `drifting_background()`
correction, same trapezoidal integral, and — critically — the same
shared 17:33:10 release anchor recovered in `01` (not each file's own
inconsistent internal time-since-release column).

```{r sr-two-station}
hydraulics %>%
  filter(event_id %in% c("SR_20231009_upstream", "SR_20231009_downstream")) %>%
  select(event_id, discharge_Ls, water_velocity_ms, t_peak_s, reach_length_m)
```

> **QAQC note — discharge.** `discharge_Ls_upstream` (4.48 L/s) now
> lines up with Kauan's own 4.30 L/s manual figure once the same
> rigorous method is applied on both sides — the earlier apparent
> mismatch was an artefact of comparing a pipeline-computed downstream
> value against a manually-computed upstream one and assuming they were
> for the same point. Downstream discharge (6.13 L/s) exceeding upstream
> by \~1.65 L/s over the \~20 m between stations reads as a genuine
> gaining reach: NaCl is conservative and the downstream curve returns
> fully to baseline (Section 6 above), so the smaller downstream
> mass-flux integral is not explained by tracer loss. It requires either
> extra water entering between the stations — diffuse groundwater or
> hyporheic upwelling is unremarkable in a karst system and would not
> necessarily show as a visible surface inflow, so this does not require
> a tributary Kauan would have seen — or a systematic offset between the
> two logger units (serials `#20835561` vs `#20835564`), which cannot be
> ruled out without an independent side-by-side calibration check.
>
> **QAQC note — velocity.** `water_velocity_ms` does not tell the same
> clean story: the upstream station, despite being *closer* to the
> release (52 m vs 72 m), peaks *later* (3110 s vs 2560 s) — 0.0167 m/s
> against 0.0281 m/s downstream. No reach hydraulics accelerates flow
> between a slower upstream reading and a faster downstream one along
> the same channel. The likely explanation is that 52 m is a
> straight-line/nominal distance, not the true flow-path length to where
> this logger actually sits: it doubles as the continuous
> metabolism-monitoring sensor, and those are conventionally sited in a
> stable pool rather than the channel thalweg, which would both delay
> and disperse the peak relative to its straight-line distance
> (consistent with its more attenuated curve — lower peak ratio — than
> downstream). `water_velocity_ms` and `stream_depth_m` for
> `SR_20231009_upstream` are flagged unreliable on this basis and should
> not feed hydraulics-dependent uses (TSM initial parameters, Section
> 9.7's areal rate); `discharge_Ls`, which depends only on mass and the
> integral and not on distance, is unaffected and is trusted.

# 7. Event-level attribute table

```{r chemistry}
chemistry <- p_raw %>%
  group_by(stream, date) %>%
  summarise(across(c(ph, temp_c, ca_mgL_up, ca_mgL_down, mg_mgL,
                     na_mgL, k_mgL, so42_mgL, cl_mgL),
                   ~ first(na.omit(.x))), .groups = "drop")
```

```{r build-events}
event_keys <- bind_rows(
  events_hobo    %>% distinct(event_id, stream, date, station, slug_label),
  nutrients_long %>% distinct(event_id, stream, date, station, slug_label)
) %>% distinct(event_id, .keep_all = TRUE)

events <- event_keys %>%
  # mass and geometry are dropped here and arrive once, via hydraulics.
  # t_peak_s is renamed, not dropped: events_hobo's version is the SpC peak
  # of the trimmed logger window (QAQC diagnostic from 01), a different
  # quantity from hydraulics' t_peak_s (the travel time behind the adopted
  # discharge source, used for water_velocity_ms). Left unrenamed, dplyr's
  # join silently suffixes both .x/.y, which the guard below catches.
  left_join(
    events_hobo %>%
      select(-stream, -date, -station, -slug_label,
             -any_of(c("nacl_mass_g", "reach_length_m", "reach_mean_width_m"))) %>%
      rename(t_peak_s_hobo_window = t_peak_s),
    by = "event_id"
  ) %>%
  left_join(chemistry,  by = c("stream", "date")) %>%
  left_join(hydraulics, by = "event_id") %>%
  left_join(
    nutrients_long %>%
      group_by(event_id) %>%
      summarise(n_grabs_nh4 = sum(solute == "NH4-N"),
                n_grabs_srp = sum(solute == "SRP"),
                nh4_background_ugL = first(na.omit(background_ugL[solute == "NH4-N"])),
                srp_background_ugL = first(na.omit(background_ugL[solute == "SRP"])),
                .groups = "drop"),
    by = "event_id"
  ) %>%
  mutate(has_logger    = event_id %in% btc_conservative$event_id,
         has_nutrients = event_id %in% nutrients_long$event_id) %>%
  arrange(stream, date)

nrow(events)                                   # esperado: 13 (11 + SR_20231009_downstream + SR_20231009_upstream)
events %>% count(event_id) %>% filter(n > 1)   # deve ser vazio
names(events) %>% str_subset("\\.(x|y)$")      # deve ser character(0)

events %>% select(event_id, nacl_mass_g, reach_length_m, reach_mean_width_m,
                  discharge_Ls, discharge_source, water_velocity_ms,
                  stream_depth_m, flag_discharge_alt, flag_discharge_invalid,
                  flag_mixing_incomplete, flag_no_baseline,
                  flag_sparse_recession, has_logger, has_nutrients) %>%
  print(n = Inf)
```

# 8. Export

```{r export}
btc_nutrients <- paired %>%
  transmute(event_id, stream, date, station, slug_label, solute,
            datetime, time_since_release_s = time_since_addition_s,
            anchor_drift_s, sample_btc, curve_limb,
            conc_ugL, background_ugL, conc_corr_ugL,
            spc_uscm_grab, spc_corr_uscm_grab, spc_uscm_logger,
            nacl_mgL_grab, nacl_mgL_grab_src, nacl_mgL_logger,
            nacl_mgL, nacl_source, gap_s) %>%
  arrange(event_id, solute, time_since_release_s)

write_csv(btc_conservative, here("data_derived", "btc_conservative.csv"))
write_csv(events,           here("data_derived", "events.csv"))

saveRDS(btc_conservative, here("data_derived", "btc_conservative.rds"))
saveRDS(events,           here("data_derived", "events.rds"))

dim(btc_nutrients); dim(btc_conservative); dim(events)
```

> **Note.** `btc_nutrients` is written in Section 9, after the
> co-precipitation columns are added, so the file on disk always carries
> the complete table.

```{r dictionary}
dict_events <- tribble(
  ~column, ~unit, ~description,
  "event_id",             "-",     "Event key; joins btc_conservative and btc_nutrients",
  "stream",               "-",     "Stream code",
  "date",                 "-",     "Event date (Brasília time)",
  "station",              "-",     "upstream/downstream for paired events, NA otherwise",
  "slug_label",           "-",     "P/N for the split RA 2023-09-06 release, NA otherwise",
  "addition_datetime",    "-",     "Tracer release time, America/Sao_Paulo",
  "nacl_mass_g",          "g",     "NaCl mass added",
  "amonium_added_g",      "g",     "Ammonium mass added",
  "phophate_added_g",     "g",     "Phosphate mass added",
  "reach_length_m",       "m",     "Reach length, authoritative survey value",
  "reach_mean_width_m",   "m",     "Mean wetted width, authoritative survey value",
  "discharge_Ls",         "L/s",   "Adopted event discharge; source given by discharge_source",
  "discharge_source",     "-",     "logger / probe / manual_excel; NA when none is valid",
  "discharge_method",     "-",     "Method behind the adopted value",
  "reason_discharge",     "-",     "Why the logger was not used, where it was not",
  "water_velocity_ms",    "m/s",   "reach_length / travel time of the adopted source",
  "stream_depth_m",       "m",     "Mean depth from continuity, adopted discharge and velocity",
  "discharge_Ls_logger",  "L/s",   "Discharge from the logger BTC (drifting baseline, trapezoidal, true peak)",
  "water_velocity_ms_logger","m/s","Mean velocity, logger estimate",
  "stream_depth_m_logger","m",     "Mean depth, logger estimate",
  "discharge_Ls_probe",   "L/s",   "Discharge from the handheld BTC (trapezoidal, end-of-rising)",
  "water_velocity_ms_probe","m/s", "Mean velocity, handheld estimate",
  "stream_depth_m_probe", "m",     "Mean depth, handheld estimate",
  "discharge_Ls_manual",  "L/s",   "Manual Excel value (fixed background, negatives retained)",
  "discharge_ratio",      "-",     "discharge_Ls_logger / discharge_Ls_probe; cross-check",
  "t_peak_s_logger",      "s",     "Time to the true SpC peak, logger",
  "t_peak_s_probe",       "s",     "End of the rising limb, handheld",
  "t_peak_s_hobo_window", "s",     "Time to the SpC peak within the trimmed logger window (01 QAQC diagnostic); distinct from t_peak_s_logger/_probe used for the adopted water_velocity_ms",
  "max_gap_s",            "s",     "Longest interval between consecutive grabs",
  "frac_area_max_gap",    "-",     "Share of the handheld trapezoidal area held by that single interval",
  "truncated_tail",       "-",     "TRUE if the logger was pulled before full baseline return",
  "flag_discharge_alt",   "-",     "TRUE when the adopted value does not come from the logger",
  "flag_discharge_invalid","-",    "TRUE when no valid discharge estimate exists",
  "flag_mixing_incomplete","-",    "TRUE when the sensor sampled the plume core, not a mixed section",
  "flag_no_baseline",     "-",     "TRUE when the window carries no clean pre-release baseline",
  "flag_weak_signal",     "-",     "TRUE when the adopted trapezoidal estimate's peak is only marginally above background noise (RA_20230906_P: ~3%)",
  "flag_sparse_recession","-",     "TRUE when one unsampled interval holds >30% of the handheld area",
  "ph",                   "-",     "Downstream water pH (event constant)",
  "temp_c",               "degC",  "Downstream water temperature (event constant)",
  "ca_mgL_up",            "mg/L",  "Calcium upstream, synoptic sample on arrival",
  "ca_mgL_down",          "mg/L",  "Calcium downstream, synoptic sample on arrival",
  "mg_mgL",               "mg/L",  "Magnesium",
  "na_mgL",               "mg/L",  "Sodium",
  "k_mgL",                "mg/L",  "Potassium",
  "so42_mgL",             "mg/L",  "Sulphate",
  "cl_mgL",               "mg/L",  "Chloride",
  "nh4_background_ugL",   "ug/L",  "Ambient NH4-N before release",
  "srp_background_ugL",   "ug/L",  "Ambient SRP before release",
  "n_grabs_nh4",          "-",     "Number of NH4-N grabs",
  "n_grabs_srp",          "-",     "Number of SRP grabs",
  "has_logger",           "-",     "TRUE if a conservative logger series exists",
  "has_nutrients",        "-",     "TRUE if nutrient grabs exist"
)

dict_btc_nutrients <- tribble(
  ~column, ~unit, ~description,
  "event_id",             "-",     "Event key",
  "solute",               "-",     "NH4-N or SRP",
  "datetime",             "-",     "Reconstructed as addition_datetime + time_since_addition_s",
  "time_since_release_s", "s",     "Elapsed time from the release anchoring this slug",
  "anchor_drift_s",       "s",     "Disagreement between spreadsheet clock and reconstructed timestamp",
  "sample_btc",           "-",     "Sample identifier along the breakthrough curve",
  "curve_limb",           "-",     "rising / peak / falling",
  "conc_ugL",             "ug/L",  "Measured nutrient concentration",
  "background_ugL",       "ug/L",  "Ambient nutrient concentration before release",
  "conc_corr_ugL",        "ug/L",  "Background-corrected nutrient concentration",
  "spc_uscm_grab",        "uS/cm", "SpC read with the handheld probe",
  "spc_corr_uscm_grab",   "uS/cm", "Background-corrected SpC from the spreadsheet (constant background)",
  "spc_uscm_logger",      "uS/cm", "Logger SpC interpolated at the grab timestamp",
  "nacl_mgL_grab",        "mg/L",  "NaCl from spc_corr_uscm_grab using NACL_SLOPE",
  "nacl_mgL_grab_src",    "mg/L",  "NaCl as it appears in the source spreadsheet",
  "nacl_mgL_logger",      "mg/L",  "Logger NaCl interpolated at the grab timestamp",
  "nacl_mgL",             "mg/L",  "Value used downstream; logger where available, grab otherwise",
  "nacl_source",          "-",     "logger_interpolated or handheld_grab",
  "gap_s",                "s",     "Distance to the nearest logger record; interpolation quality"
)

write_csv(dict_events, here("data_derived", "dictionary", "dict_events.csv"))
dict_events
```

# 9. CaCO₃ co-precipitation correction for SRP

## 9.1 Why the correction is needed

In alkaline karst streams part of the dissolved inorganic phosphorus is
removed abiotically, adsorbed onto calcite as it precipitates, rather
than assimilated by organisms. In the breakthrough curve that removal is
indistinguishable from biological uptake: both make phosphorus disappear
faster than the conservative tracer. Left uncorrected it inflates every
phosphorus uptake metric, and the inflation is largest exactly where
calcite precipitation is strongest — a systematic bias correlated with
the variable of interest, not noise.

## 9.2 What the correction does

Following House (1986, 1990), the co-precipitated P at each grab is

$$ n_{P,\mathrm{cop}} = \Delta n_{\mathrm{Ca}}\,(\sigma\,N_A\,d)\,h $$

where $\Delta n_{\mathrm{Ca}}$ is the net loss of dissolved Ca²⁺ across
the reach (mol L⁻¹), $\sigma$ the maximum surface density of P on
precipitating calcite (5.5 × 10⁻⁷ mol m⁻²), $d$ the molecular surface
area of calcite (2.01 × 10⁻¹⁹ m² molecule⁻¹), $N_A$ Avogadro's number,
and $h$ the fraction of dissolved phosphate liable to sorb, from the
two-component Langmuir model

$$ h = \frac{K_1 a_{\mathrm{PO}_4^{3-}} + K_2 a_{\mathrm{HPO}_4^{2-}}}
             {1 + K_1 a_{\mathrm{PO}_4^{3-}} + K_2 a_{\mathrm{HPO}_4^{2-}}} $$

with activities from PHREEQC and the adsorption constants
temperature-dependent through Arrhenius expressions,

$$ K_1(T) = A_1 \exp\!\left(\frac{E_1}{RT}\right), \qquad
   K_2(T) = A_2 \exp\!\left(\frac{E_2}{RT}\right) $$

where $A_1 = 0.6915$, $A_2 = 4.361 \times 10^{-9}$, $E_1 = 18.2$ kJ
mol⁻¹, $E_2 = 42.6$ kJ mol⁻¹, $R$ = 8.314 J mol⁻¹ K⁻¹ and $T$ is
downstream water temperature in kelvin.

**Where each term comes from.** The activities are computed from the
**downstream** solution — the water passing the sampling station is what
must be speciated — using downstream pH, temperature and major ions,
with `ca_mgL_down` rather than any upstream-downstream mean.
$\Delta n_{\mathrm{Ca}}$ comes from a synoptic pair: upstream and
downstream calcium sampled on arrival at the reach, before the nutrient
release. It measures the calcite precipitation the reach was sustaining
under ambient conditions, so it is one value per event, and the
within-event variation in $n_{P,\mathrm{cop}}$ comes entirely from $h$,
which tracks phosphate speciation as the slug passes. This follows from
the design: a single ambient pair characterises the reach, and the model
scales it by how much phosphate is present at each instant.

Two consequences are stated rather than worked around. The calcium pair
predates the release, so it cannot capture any change in calcite
dynamics caused by the addition itself. And being a single pair, the
correction carries no within-event uncertainty — its uncertainty is
entirely between events.

**Data provenance note.** The `ph` and `temp_c` columns of the nutrient
spreadsheets originally carried the upstream reading for some events, a
field-sheet transcription error corrected before the speciation was run.
Since 0.1 pH unit and 1 °C both shift $K_1$, $K_2$ and the PO₄³⁻/HPO₄²⁻
partition measurably, the guard in `R/phreeqc_speciation.R` re-checks
every event against the synoptic upstream table and warns if any still
matches the upstream value exactly.

## 9.3 How it enters the pipeline, and why not as an overwrite

Under TASCC the correction feeds a per-sample regression of
$S_{w\text{-add}}$ against the geometric mean concentration, so each
grab yields its own uptake length and a corrected point simply replaces
an uncorrected one. Under inverse transient-storage modelling the
corrected series **is** the observation vector: the whole curve is
fitted at once, and a correction applied unevenly along it changes the
shape the model is matching, not just the value at a point. Overwriting
would make the two analyses non-comparable and irreversible.

The correction is therefore added as parallel columns and
`conc_corr_ugL` is never modified, so both series can be run through the
inverse model and the difference attributed to the abiotic pathway.

**Naming.** The suffix `_corr` already means "background-corrected"
throughout this pipeline (`conc_corr_ugL`, `spc_corr_uscm`,
`spc_corr_uscm_grab`). The co-precipitation column extends that same
series rather than introducing a second meaning, so it is
`conc_corr_coprec_ugL`: the background-corrected concentration, plus the
co-precipitated fraction. It is `NA` for NH4-N, like every other
SRP-specific column in this long-format table.

## 9.4 Calcium change along the reach

The House model needs to know how much dissolved calcium the reach
removed from the water. That comes from `calcium_reach_change.csv`: two
water samples per event, one at the top of the reach and one at the
bottom, both collected on arrival before anything was released. The same
sampling gave pH and specific conductance at both ends.

Because the two samples are taken at the same time, the difference
between them is a difference in space — what the reach did to the water
passing through it — and not a change over the course of the day. This
is the file the pipeline uses for calcium, pH and conductance; it
replaces the values carried in the nutrient spreadsheets so these
quantities have one source.

The convention in the file is downstream minus upstream, so **negative
means calcium was lost along the reach**, which is what calcite
precipitation does. The House model of Section 9.2 wants that loss as a
positive number, so `delta_ca_mgL` in Section 9.5 is the negation of
this column.

### The three ratios

Each variable is expressed as a log response ratio: the natural log of
the downstream value over the upstream value.

$$ \mathrm{LRR}_{X} = \ln\!\frac{X_{\mathrm{down}}}{X_{\mathrm{up}}} $$

Zero means no change, negative means the reach removed it. The log form
is used because it makes losses and gains symmetric — halving and
doubling are equal and opposite — and because calcium (tens of mg L⁻¹)
and conductance (hundreds of µS cm⁻¹) then sit on the same scale.

pH is already a logarithm, so a ratio of pH values means nothing
physical. The ratio is taken on hydrogen ion activity instead:

$$ \mathrm{LRR}_{\mathrm{H}^+} =
   -\ln(10)\,\bigl(\mathrm{pH}_{\mathrm{down}} - \mathrm{pH}_{\mathrm{up}}\bigr) $$

Negative means the water alkalinised.

Calcium and conductance together separate two mechanisms. Water entering
the reach from elsewhere dilutes everything in proportion, moving both
ratios equally. Only selective removal moves calcium further than
conductance, so `lrr_ca_excess`, the difference between the two, is the
part of the calcium change that dilution cannot account for.

```{r coprec-geochem-check}
reach_pair <- read_csv(here("data", "field", "calcium_reach_change.csv"),
                       show_col_types = FALSE) %>%
  mutate(date = ymd(as.character(date))) %>%
  rename(ph_up = ph_upstream, ph_down = ph_dowstream,
         spc_uscm_up = spc_upstream, spc_uscm_down = spc_downstream,
         ca_mgL_up = ca_upstream, ca_mgL_down = ca_downstream)

geochem_check <- reach_pair %>%
  mutate(
    # file convention: downstream - upstream, so negative = calcium lost
    ca_mass_trans_mgL = ca_mgL_down - ca_mgL_up,

    lrr_ca  = log(ca_mgL_down / ca_mgL_up),
    lrr_spc = log(spc_uscm_down / spc_uscm_up),
    lrr_h   = -log(10) * (ph_down - ph_up),

    # calcium change beyond what dilution explains
    lrr_ca_excess = lrr_ca - lrr_spc,

    ca_change_pct = 100 * (exp(lrr_ca) - 1),

    # is this one of the events analysed in this study?
    in_study = paste(stream, format(date, "%Y%m%d")) %in%
               paste(events$stream, format(events$date, "%Y%m%d"))
  ) %>%
  arrange(stream, date)

geochem_check %>%
  select(stream, date, in_study, ca_mgL_up, ca_mgL_down, ca_mass_trans_mgL,
         ca_change_pct, lrr_ca, lrr_spc, lrr_ca_excess, lrr_h) %>%
  print(n = Inf)
```

### Per stream

Each stream was sampled on more than one date. Averaging the log ratios
is the correct way to average ratios — the mean of the logs is the log
of the geometric mean, whereas averaging percentages is not meaningful.

The spread between dates is reported alongside the mean, because it is
the more informative number here. Calcium transfer varies within a
stream, between streams, and from one day to the next, so a stream mean
is only a useful summary when the dates agree on direction. Where they
do not, `same_direction` is `FALSE` and the mean should not be read as a
property of the stream.

```{r coprec-geochem-by-stream}
geochem_by_stream <- geochem_check %>%
  group_by(stream) %>%
  summarise(
    n_dates = n(),
    dates   = paste(format(date, "%d/%m"), collapse = ", "),

    ca_delta_mean = mean(ca_mass_trans_mgL),
    ca_delta_min  = min(ca_mass_trans_mgL),
    ca_delta_max  = max(ca_mass_trans_mgL),

    lrr_ca_mean = mean(lrr_ca),
    lrr_ca_sd   = if (n() > 1) sd(lrr_ca) else NA_real_,
    lrr_ca_excess_mean = mean(lrr_ca_excess, na.rm = TRUE),
    lrr_h_mean  = mean(lrr_h, na.rm = TRUE),

    same_direction = n_distinct(sign(ca_mass_trans_mgL)) == 1,
    .groups = "drop"
  ) %>%
  mutate(ca_change_pct_mean = 100 * (exp(lrr_ca_mean) - 1)) %>%
  arrange(lrr_ca_mean)

geochem_by_stream %>% print(n = Inf)
```

```{r coprec-geochem-plot, fig.height=4.5, fig.width=11}
ord <- geochem_check %>%
  mutate(event = paste(stream, format(date, "%d/%m"))) %>%
  arrange(lrr_ca) %>%
  pull(event)

geochem_check %>%
  mutate(event = factor(paste(stream, format(date, "%d/%m")), levels = ord)) %>%
  select(event, stream, lrr_ca, lrr_spc, lrr_h) %>%
  pivot_longer(starts_with("lrr_"), names_to = "series", values_to = "lrr") %>%
  filter(!is.na(lrr)) %>%
  mutate(series = factor(series,
    levels = c("lrr_ca", "lrr_spc", "lrr_h"),
    labels = c("Ca", "SpC", "pH"))) %>%
  ggplot(aes(lrr, event, colour = stream)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4) +
  geom_segment(aes(x = 0, xend = lrr, yend = event), linewidth = 0.5) +
  geom_point(size = 3) +
  facet_wrap(~ series, nrow = 1, scales = "free_x") +
  labs(x = "log response ratio (ln down / up)", y = NULL) +
  cowplot::theme_cowplot(font_size = 14) +
  theme(legend.position = "none",
        panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.3))
```

> **QAQC note.** Negative calcium change means the reach removed
> calcium, which precipitation produces; positive means it gained
> calcium, which precipitation cannot explain. Most events show a small
> gain, so co-precipitation is not a system-wide process in this
> dataset: the correction has a material effect only where calcium was
> clearly removed, and no effect at all where it was gained.
>
> Calcium was determined by EDTA titration at a commercial laboratory
> and reported to 0.1 mg L⁻¹. Several events differ by exactly 0.4 mg
> L⁻¹, which is close to the reporting precision, so those changes are
> recorded but carry little weight. The uncertainty is quantified in
> Section 9.6.
>
> `lrr_spc` and `lrr_h` are `NA` where the upstream reading is missing
> from the file; `lrr_ca` is defined for every event. \`\`\`

## 9.5 Application

```{r coprec-constants}
# SIGMA_P: House (1990) reports several sigma estimates, and they cluster by
# solution type rather than scattering randomly -- field/river waters run
# 0.055-0.095 umol/m2 (River Frome, River Windrush, Waterston chalk
# aquifer), while idealised synthetic Ca(HCO3)2 solutions run 0.13-0.22
# umol/m2. This is a karst STREAM, not a synthetic solution, so the field
# cluster is the physically appropriate class. 5.5e-8 mol/m2 is the River
# Frome value (House et al. 1986a; the "river water" case House 1990
# itself highlights) -- the same value and reasoning Cohen et al. (2013,
# Ecol. Monogr. 83:2) used for the Ichetucknee River, a spring-fed
# karst/calcite stream directly analogous to this system, explicitly
# preferring the field value because natural water carries competing
# solutes (Mg, Fe, DOM) that reduce the effective sorption area, and
# because their own sensitivity check found the lab-range value inflates
# the correction ~47x and yields implausible C:P ratios. See Section 9.8
# for the corresponding sensitivity range in this pipeline.
SIGMA_P   <- 5.5e-8        # mol/m2 -- River Frome (House 1990 / Cohen et al. 2013)
D_CALCITE <- 2.01e-19      # m2/molecule, molecular surface area of calcite
N_AVOG    <- 6.02214076e23 # molecules/mol
R_GAS     <- 8.314         # J/mol/K
A1 <- 0.6915  ; E1 <- 18.2e3   # J/mol
A2 <- 4.361e-9; E2 <- 42.6e3   # J/mol

M_P  <- 30.973762   # g/mol
M_CA <- 40.078      # g/mol

h_langmuir2 <- function(a_PO4, a_HPO4, K1, K2) {
  num <- K1 * a_PO4 + K2 * a_HPO4
  num / (1 + num)
}
```

```{r coprec-apply}
act_path <- here("data_derived", "phreeqc_activities.csv")

if (!file.exists(act_path)) {
  warning("phreeqc_activities.csv nao encontrado: rode R/phreeqc_speciation.R. ",
          "As colunas de coprecipitacao ficarao NA.")
  coprec <- tibble(stream = character(), date = as.Date(character()),
                   sample_btc = numeric())
} else {
  # date is written as a bare integer (20231008) by the speciation script, so
  # it must be parsed with ymd(); as.Date() on a number silently yields a date
  # in year 57345 and every join returns NA.
  coprec <- read_csv(act_path, show_col_types = FALSE) %>%
    mutate(date = ymd(as.character(date))) %>%
    left_join(
      p_raw %>% select(stream, date, sample_btc, temp_c, ca_mgL_up, ca_mgL_down),
      by = c("stream", "date", "sample_btc")
    ) %>%
    mutate(
      T_K = temp_c + 273.15,
      k1_langmuir = A1 * exp(E1 / (R_GAS * T_K)),
      k2_langmuir = A2 * exp(E2 / (R_GAS * T_K)),
      # House's K1/K2 (Eq. 7-8) and the Langmuir isotherm (Eq. 9, House
      # 1990) are defined for [PO4^3-]/[HPO4^2-] in umol/dm^3 -- House
      # states this explicitly under Eq. 8 ("the concentrations of the
      # phosphate and hydrogenphosphate ions are expressed in units of
      # umol dm^-3"). PHREEQC's -activities output is on a mol/L numeric
      # scale, six orders of magnitude smaller; passing it to
      # h_langmuir2() unconverted collapses h_adsorbed toward ~0
      # regardless of how saturated the solution actually is -- this is
      # the second, larger reason the coprecipitation correction still
      # looked negligible after the SIGMA_P and major-ion mg/L fixes.
      h_adsorbed  = h_langmuir2(a_PO4 * 1e6, a_HPO4 * 1e6, k1_langmuir, k2_langmuir),

      # synoptic pair on arrival; positive = calcium lost across the reach
      delta_ca_mgL  = ca_mgL_up - ca_mgL_down,
      delta_ca_molL = delta_ca_mgL / (1000 * M_CA),

      p_coprec_molL = delta_ca_molL * (SIGMA_P * N_AVOG * D_CALCITE) * h_adsorbed,
      p_coprec_ugL  = p_coprec_molL * M_P * 1e6,

      flag_delta_ca_neg = delta_ca_mgL < 0
    ) %>%
    select(stream, date, sample_btc, a_PO4, a_HPO4, h_adsorbed,
           k1_langmuir, k2_langmuir, delta_ca_mgL, p_coprec_ugL,
           flag_delta_ca_neg)
}

# guard: the join must match. If this returns 0 rows the date types disagree.
cat("eventos com atividades pareadas:",
    nrow(distinct(coprec, stream, date)), "\n")

# Adding p_coprec to the background-corrected concentration reconstructs the
# SRP that would have been present without abiotic removal:
#   (measured + coprecipitated) - background
# Background subtraction and this addition commute, so order is immaterial.
btc_nutrients <- btc_nutrients %>%
  select(-any_of(c("a_PO4", "a_HPO4", "h_adsorbed", "k1_langmuir",
                   "k2_langmuir", "delta_ca_mgL", "p_coprec_ugL",
                   "flag_delta_ca_neg", "conc_corr_coprec_ugL"))) %>%
  left_join(coprec, by = c("stream", "date", "sample_btc")) %>%
  mutate(
    across(c(a_PO4, a_HPO4, h_adsorbed, k1_langmuir, k2_langmuir,
             delta_ca_mgL, p_coprec_ugL),
           ~ if_else(solute == "SRP", .x, NA_real_)),
    flag_delta_ca_neg = if_else(solute == "SRP", flag_delta_ca_neg, NA),
    conc_corr_coprec_ugL = if_else(solute == "SRP",
                                   conc_corr_ugL + p_coprec_ugL,
                                   NA_real_)
  )

btc_nutrients %>%
  filter(solute == "SRP") %>%
  group_by(event_id) %>%
  summarise(
    n = n(),
    delta_ca_mgL = first(delta_ca_mgL),
    h_median     = median(h_adsorbed, na.rm = TRUE),
    coprec_median_ugL = median(p_coprec_ugL, na.rm = TRUE),
    frac_of_srp  = median(p_coprec_ugL / conc_corr_ugL, na.rm = TRUE),
    delta_ca_neg = first(flag_delta_ca_neg),
    .groups = "drop"
  ) %>%
  arrange(desc(delta_ca_mgL)) %>%
  print(n = Inf)
```

> **QAQC note.** `frac_of_srp` is the size of the correction relative to
> the signal being corrected, and it decides how much any of this
> matters. Where it is small, the treatment of ΔCa \< 0 cannot change a
> conclusion; where it approaches unity, the corrected series is largely
> model output rather than measurement, and the inverse fit should be
> reported for both series.
>
> Where `delta_ca_neg` is TRUE the reach gained calcium: calcite
> dissolved rather than precipitated. The House model returns a negative
> $n_{P,\mathrm{cop}}$ there, so the correction subtracts phosphorus,
> which is outside the direction the model was parameterised for. No
> precedent for this case exists in the BTC literature, so these events
> are flagged and carried forward, and robustness of the main
> conclusions is demonstrated across alternative treatments of them
> rather than one treatment being asserted as correct.

## 9.6 Calcium change: measurement uncertainty

`ca_mass_trans_mgL` (§9.4) is a difference of two independent EDTA
titrations, upstream and downstream. Whether an observed difference
reflects real transport or just analytical noise depends on how large it
is relative to that noise, which this section quantifies rather than
asserts.

The laboratory reports each titration to 0.1 mg/L, but several events
differ by exactly that increment, which is a sign the *reporting*
precision is finer than the *analytical* precision. Standard Methods
3500-Ca gives two different figures for EDTA titrimetric calcium: an
interlaboratory relative standard deviation of 9.2 %, and a relative
error of 1.9 %. The two are not interchangeable — the first is a spread
across labs and operators (a standard deviation), the second is a bias
against a reference value. A standard deviation is what a z-score needs,
so **`CA_RSD` is set to 9.2 %**: the only one of the two that is a
standard deviation, and the conservative (larger) choice.

Each of the two titrations (upstream, downstream) is treated as an
independent measurement with relative standard deviation `CA_RSD` of its
own concentration. For two independent errors of comparable relative
size, the standard deviation of their difference is

$$ \sigma_{\Delta\mathrm{Ca}} = \sqrt{2}\;\mathrm{CA\_RSD}\;\overline{\mathrm{Ca}} $$

where $\overline{\mathrm{Ca}}$ is the mean of the upstream and
downstream concentrations (used as the reference level since the true
concentration is unknown) and the $\sqrt{2}$ comes from adding two
equal-variance, independent errors in quadrature. The observed transfer
is then expressed in units of its own uncertainty,
$z = \Delta\mathrm{Ca}_{\mathrm{mass\ trans}} / \sigma_{\Delta\mathrm{Ca}}$;
`flag_ca_resolved` marks events where $|z| > 1$, i.e. where the observed
change exceeds one analytical standard deviation. No event is excluded
by this — every row keeps its value and its flag, and the areal rate of
§9.7 is reported for both resolved and unresolved events alike.

```{r ca-uncertainty}
CA_RSD <- 0.092   # Standard Methods 3500-Ca, EDTA titrimetric; interlaboratory
                  # RSD. NOT the method's 1.9% relative error, which is a bias
                  # against a reference, not a standard deviation.

geochem_check <- geochem_check %>%
  mutate(
    ca_mean_mgL      = (ca_mgL_up + ca_mgL_down) / 2,
    ca_delta_sd      = sqrt(2) * CA_RSD * ca_mean_mgL,
    ca_delta_z       = ca_mass_trans_mgL / ca_delta_sd,
    flag_ca_resolved = abs(ca_delta_z) > 1
  )

geochem_check %>%
  select(stream, date, ca_mass_trans_mgL, ca_mean_mgL, ca_delta_sd,
         ca_delta_z, flag_ca_resolved) %>%
  arrange(abs(ca_delta_z)) %>%
  print(n = Inf)
```

> **QAQC note.** `r sum(geochem_check$flag_ca_resolved)` of
> `r nrow(geochem_check)` event-dates have `|ca_delta_z| > 1`. Given the
> key finding of §9.4 — most reaches show a small calcium *gain* rather
> than a loss — this uncertainty check matters most for the gains: a
> change of +0.4 mg/L sitting at $|z|<1$ is not distinguishable from
> analytical noise, whereas SR 2023-10-11 (−10.4 mg/L) and CD 2023-09-11
> (−5.6 mg/L) are unambiguous regardless of this correction.

## 9.7 Areal calcium precipitation rate

A concentration difference alone is not comparable between reaches: 1
mg/L lost over a 52 m reach (CD) reflects a far more intense process
than the same 1 mg/L over 408 m (RS). Normalising by reach length and
residence time, and by wetted area, gives a rate that **is** comparable:

$$ r_A = \frac{\Delta[\mathrm{Ca}]}{L/v}\cdot 10^3 \cdot z
   \qquad\text{(route 1: BTC hydraulics)} $$

$$ r_A = \frac{Q\,\Delta[\mathrm{Ca}]}{L\,w}
   \qquad\text{(route 2: campaign discharge)} $$

Both reduce to the same quantity under continuity ($Q = v\,w\,z$) and
are computed side by side as a mutual check: the $10^3$ in route 1
converts $\Delta[\mathrm{Ca}]$ from mg/L to mg/m³ so both routes come
out in the same units, mg m⁻² s⁻¹. $\Delta[\mathrm{Ca}]$ is
`ca_mass_trans_mgL` negated (§9.4 convention: positive = calcium lost,
matching the sign `r_A` is meant to carry), $L$ is the calcium reach
length of `calcium_reach_change.csv` (not necessarily the BTC slug reach
— the two differ for several events, e.g. CB 2023-08-31 at 252 m against
162 m for the BTC reach), and $w$ is the fixed per-stream width of
`data/field/streams_width.csv`.

The two routes deliberately draw on **different** hydraulic estimates.
Route 1 uses $v$ and $z$ from the adopted BTC hydraulics (`events`) of
that stream-date — internally self-consistent (both derived from the
same adopted discharge) but only defined where a single, unambiguous BTC
event exists for the date. Two calcium dates (RA 2023-09-06, split into
P/N slugs; RA 2023-10-05, split into upstream/downstream stations) have
no single reach-level $v,z$ and are left `NA` in route 1, flagged
`flag_no_btc_v`; route 2 is unaffected since it does not depend on a BTC
event at all. Route 2 uses $Q$ from `discharge_for_calcium.csv`, per
Kauan's decision to use the probe/handheld-derived discharge there since
the calcium grabs and the discharge gaugings share the same manual
sampling scheme. Where the two routes disagree, the disagreement is
attributed to the hydraulic terms ($v$, $z$, $Q$, or $w$), not to the
chemistry, since both start from the same $\Delta[\mathrm{Ca}]$ and $L$.

`streams_width.csv` has no entry for RC, so both of RC's event-dates are
flagged `flag_no_width` and route 2 (and the width-dependent parts of
the mutual check) is `NA` there — not silently substituted.

Because $r_A$ is linear in $Q$, the spread across candidate discharges
at a site (`Q_min_ponto`, `Q_max_ponto` from `qaqc_discharge_records.R`,
and `Q_alt` for RA 2023-08-04's alternative mass reading) **is** the
spread of the rate, so it is reported alongside the point estimate
rather than folded into a single number.

```{r ca-areal-rate}
# Route 1 needs one unambiguous velocity and depth per stream-date, which
# only exists for single (non-split) BTC events.
btc_hydraulics_single <- events %>%
  filter(is.na(station), is.na(slug_label)) %>%
  transmute(stream, date, v_btc_ms = water_velocity_ms, z_btc_m = stream_depth_m)

streams_width <- read_csv(here("data", "field", "streams_width.csv"),
                          show_col_types = FALSE)

discharge_for_calcium <- read_csv(here("data_derived", "discharge_for_calcium.csv"),
                                  show_col_types = FALSE) %>%
  transmute(stream = Site, date = date_ca, Q_calcium_Ls = Q_Ls,
            Q_min_ponto, Q_max_ponto, Q_alt, flag_sem_vazao)

areal_ca <- geochem_check %>%
  transmute(stream, date, reach_length_m, ca_mass_trans_mgL,
            ca_loss_mgL = -ca_mass_trans_mgL) %>%   # House convention: positive = lost
  left_join(btc_hydraulics_single, by = c("stream", "date")) %>%
  left_join(streams_width,         by = "stream") %>%
  left_join(discharge_for_calcium, by = c("stream", "date")) %>%
  mutate(
    flag_no_width = is.na(mean_width_m),
    flag_no_btc_v = is.na(v_btc_ms),
    flag_no_q     = is.na(Q_calcium_Ls) | coalesce(flag_sem_vazao, TRUE),

    r_areal_btc_mgm2s      = (ca_loss_mgL / (reach_length_m / v_btc_ms)) * 1000 * z_btc_m,
    r_areal_campaign_mgm2s = (Q_calcium_Ls * ca_loss_mgL) / (reach_length_m * mean_width_m),
    r_areal_Qmin_mgm2s     = (Q_min_ponto  * ca_loss_mgL) / (reach_length_m * mean_width_m),
    r_areal_Qmax_mgm2s     = (Q_max_ponto  * ca_loss_mgL) / (reach_length_m * mean_width_m),
    r_areal_Qalt_mgm2s     = (Q_alt        * ca_loss_mgL) / (reach_length_m * mean_width_m),

    # per-day convenience unit alongside the SI (mg m-2 s-1) value
    r_areal_campaign_mgm2d = r_areal_campaign_mgm2s * 86400,

    route_disagreement_pct = 100 * (r_areal_btc_mgm2s - r_areal_campaign_mgm2s) /
                              r_areal_campaign_mgm2s
  ) %>%
  arrange(stream, date)

areal_ca %>%
  select(stream, date, ca_loss_mgL, reach_length_m, mean_width_m,
         r_areal_btc_mgm2s, r_areal_campaign_mgm2s, r_areal_campaign_mgm2d,
         route_disagreement_pct, flag_no_width, flag_no_btc_v, flag_no_q) %>%
  print(n = Inf)

cat("\n--- espalhamento da vazao (rota da campanha) ---\n")
areal_ca %>%
  select(stream, date, r_areal_Qmin_mgm2s, r_areal_campaign_mgm2s,
         r_areal_Qmax_mgm2s, r_areal_Qalt_mgm2s) %>%
  print(n = Inf)

write_csv(areal_ca, here("data_derived", "calcium_areal_rate.csv"))

dict_ca_uncertainty <- tribble(
  ~column, ~unit, ~description,
  "ca_mean_mgL",     "mg/L", "Mean of ca_mgL_up and ca_mgL_down; reference level for the relative-uncertainty propagation",
  "ca_delta_sd",      "mg/L", "sqrt(2)*CA_RSD*ca_mean_mgL; propagated SD of ca_mass_trans_mgL from two independent EDTA titrations",
  "ca_delta_z",       "-",    "ca_mass_trans_mgL / ca_delta_sd",
  "flag_ca_resolved", "-",    "TRUE when |ca_delta_z| > 1, i.e. the observed calcium change exceeds one analytical standard deviation"
)

dict_areal_ca <- tribble(
  ~column, ~unit, ~description,
  "ca_loss_mgL",             "mg/L",        "-ca_mass_trans_mgL; positive = calcium lost (House convention, matches Section 9.4)",
  "v_btc_ms",                 "m/s",         "water_velocity_ms from events, single (non-split) BTC event on that stream-date only",
  "z_btc_m",                  "m",           "stream_depth_m from events, same restriction as v_btc_ms",
  "mean_width_m",             "m",           "Fixed per-stream reach width, data/field/streams_width.csv; no entry for RC",
  "Q_calcium_Ls",              "L/s",         "Discharge assigned to this calcium sampling, discharge_for_calcium.csv (probe/handheld scheme)",
  "r_areal_btc_mgm2s",         "mg m-2 s-1", "Areal calcium rate, route 1 (BTC v, z); NA where v_btc_ms/z_btc_m undefined",
  "r_areal_campaign_mgm2s",    "mg m-2 s-1", "Areal calcium rate, route 2 (campaign Q); NA where width or Q undefined",
  "r_areal_campaign_mgm2d",    "mg m-2 d-1", "r_areal_campaign_mgm2s * 86400; reporting convenience only",
  "r_areal_Qmin_mgm2s",        "mg m-2 s-1", "Route 2 at Q_min_ponto (lowest discharge on record at that site)",
  "r_areal_Qmax_mgm2s",        "mg m-2 s-1", "Route 2 at Q_max_ponto (highest discharge on record at that site)",
  "r_areal_Qalt_mgm2s",        "mg m-2 s-1", "Route 2 at Q_alt; only defined for RA 2023-08-04 (alternative 12 kg mass reading)",
  "route_disagreement_pct",   "%",           "100*(r_areal_btc - r_areal_campaign)/r_areal_campaign; large values flag hydraulic, not chemical, disagreement",
  "flag_no_width",             "-",          "TRUE when streams_width.csv has no entry for this stream (currently RC)",
  "flag_no_btc_v",             "-",          "TRUE when no single BTC event exists for this stream-date (split events or no logger event that day)",
  "flag_no_q",                 "-",          "TRUE when discharge_for_calcium.csv carries no usable Q for this date"
)

write_csv(bind_rows(dict_ca_uncertainty, dict_areal_ca),
          here("data_derived", "dictionary", "dict_geochem_extra.csv"))
```

> **QAQC note.** Where both routes are defined, `route_disagreement_pct`
> is the diagnostic — a large disagreement flags the hydraulics, not a
> chemistry problem, since both routes share the same
> $\Delta[\mathrm{Ca}]$ and $L$. `r_areal_campaign_mgm2d` is reported
> purely for interpretability (a manuscript reads more naturally in mg
> m⁻² d⁻¹ than mg m⁻² s⁻¹); the SI value is what all comparisons in this
> document use. This is new analysis, not yet cross-checked against
> literature co-precipitation rates — treat the point estimates as
> provisional until that comparison is done.

```{r coprec-export}
write_csv(btc_nutrients, here("data_derived", "btc_nutrients.csv"))
saveRDS(btc_nutrients,   here("data_derived", "btc_nutrients.rds"))
write_csv(geochem_check, here("data_derived", "geochem_check.csv"))

dict_coprec <- tribble(
  ~column, ~unit, ~description,
  "a_PO4",               "-",     "PO4(3-) activity, PHREEQC, downstream solution",
  "a_HPO4",              "-",     "HPO4(2-) activity, PHREEQC, downstream solution",
  "k1_langmuir",         "-",     "Two-component Langmuir constant for PO4(3-), Arrhenius in temp_c",
  "k2_langmuir",         "-",     "Two-component Langmuir constant for HPO4(2-), Arrhenius in temp_c",
  "h_adsorbed",          "-",     "Fraction of dissolved phosphate liable to sorb onto new calcite",
  "delta_ca_mgL",        "mg/L",  "ca_mgL_up - ca_mgL_down, synoptic pair on arrival; positive = Ca lost",
  "p_coprec_ugL",        "ug/L",  "Inorganic P co-precipitated with calcite, House (1990)",
  "conc_corr_coprec_ugL","ug/L",  "conc_corr_ugL + p_coprec_ugL; SRP without abiotic removal. NA for NH4-N",
  "flag_delta_ca_neg",   "-",     "TRUE when the reach gained calcium; model not parameterised for this direction"
)

write_csv(bind_rows(dict_btc_nutrients, dict_coprec),
          here("data_derived", "dictionary", "dict_btc_nutrients.csv"))

dim(btc_nutrients)
dict_coprec
```

## 9.8 Sensitivity of the coprecipitation correction to σ

`SIGMA_P` is the single most consequential constant in this correction —
`p_coprec_ugL` is exactly linear in it — and the published estimates
cluster into two physically distinct classes rather than forming one
continuous range: field/river waters (0.055–0.095 µmol/m², River Frome,
River Windrush, the Waterston chalk aquifer) and idealised synthetic
Ca(HCO₃)₂ solutions (0.13–0.22 µmol/m²), which run 2–4× higher because
they lack the competing solutes (Mg, Fe, dissolved organic matter) that
occupy sorption sites in real water. `SIGMA_P` (§9.5) adopts the low end
of the field cluster — River Frome, 5.5e-8 mol/m² — following Cohen et
al. (2013)'s identical choice for a directly analogous karst stream;
this section carries the rest of the field cluster and the laboratory
ceiling forward as an explicit envelope rather than asserting the point
estimate as the only possibility.

```{r coprec-sigma-sensitivity}
# p_coprec_molL was built as delta_ca_molL * (SIGMA_P * N_AVOG * D_CALCITE) * h_adsorbed,
# so for fixed delta_ca and h it is exactly proportional to SIGMA_P: rescaling
# the already-computed p_coprec_ugL is equivalent to, and cheaper than,
# rebuilding the join for each candidate sigma.
SIGMA_SENS <- c(
  field_low   = 5.5e-8,  # River Frome (House et al. 1986a) -- ADOPTED, = SIGMA_P
  field_high  = 9.5e-8,  # Waterston chalk aquifer (House et al. 1986a), field cluster upper end
  lab_ceiling = 1.5e-7   # synthetic Ca(HCO3)2 (House & Donaldson 1986); Cohen et al. (2013)'s
                         # laboratory upper-bound comparison value
)
stopifnot(isTRUE(all.equal(unname(SIGMA_SENS["field_low"]), SIGMA_P)))

coprec_sigma_sens <- btc_nutrients %>%
  filter(solute == "SRP", !is.na(p_coprec_ugL)) %>%
  select(event_id, sample_btc, p_coprec_ugL, conc_corr_ugL) %>%
  crossing(sigma_label = names(SIGMA_SENS)) %>%
  mutate(
    sigma_molm2       = SIGMA_SENS[sigma_label],
    p_coprec_ugL_sens = p_coprec_ugL * (sigma_molm2 / SIGMA_P),
    frac_of_srp_sens  = p_coprec_ugL_sens / conc_corr_ugL
  )

coprec_sigma_sens %>%
  group_by(event_id, sigma_label) %>%
  summarise(
    sigma_umolm2        = first(sigma_molm2) * 1e6,
    coprec_median_ugL   = median(p_coprec_ugL_sens, na.rm = TRUE),
    frac_of_srp_median  = median(frac_of_srp_sens,  na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(id_cols = event_id, names_from = sigma_label,
              values_from = c(coprec_median_ugL, frac_of_srp_median)) %>%
  arrange(event_id) %>%
  print(n = Inf)

write_csv(coprec_sigma_sens, here("data_derived", "coprec_sigma_sensitivity.csv"))

dict_coprec_sigma_sens <- tribble(
  ~column, ~unit, ~description,
  "sigma_label",       "-",      "field_low (adopted, River Frome) / field_high (Waterston aquifer) / lab_ceiling (synthetic solution)",
  "sigma_molm2",        "mol/m2", "SIGMA_P value used for this row, from SIGMA_SENS",
  "p_coprec_ugL_sens",  "ug/L",   "p_coprec_ugL rescaled to sigma_molm2; linear rescaling of the SIGMA_P-based value",
  "frac_of_srp_sens",   "-",      "p_coprec_ugL_sens / conc_corr_ugL, at this sigma"
)
write_csv(dict_coprec_sigma_sens,
          here("data_derived", "dictionary", "dict_coprec_sigma_sensitivity.csv"))
```

> **QAQC note.** This is the same check Cohen et al. (2013) ran on the
> Ichetucknee: moving from the field value to the laboratory ceiling
> should not be treated as a free parameter choice, because at the
> laboratory end `frac_of_srp` can approach or exceed unity — the point
> where the corrected series stops being a correction and becomes mostly
> model output (§9.5's QAQC note). Where `frac_of_srp_sens` stays small
> across the whole field-to-lab envelope, the choice of `SIGMA_P` cannot
> change a conclusion; where it diverges sharply between `field_low` and
> `lab_ceiling`, that divergence — not the point estimate — is what
> belongs in the manuscript, alongside the physical argument (§9.5) for
> why the field cluster is the appropriate class for a stream rather
> than a synthetic solution.

## 9.9 Event-level mass balance: does the correction respect the injected P mass?

`frac_of_srp` (§9.5) is a per-grab, concentration-domain ratio: at each
sampled instant, how large the correction is relative to the SRP
measured at that instant. The `median(frac_of_srp)` printed there is the
size of a *typical* correction, not the size of what the event actually
lost — it weighs every grab equally regardless of when it was taken or
how much water carried it, and `h_adsorbed` tracks phosphate speciation
(§9.2), so `p_coprec_ugL` rises and falls with SRP across the curve
while most events sample the tail more densely than the peak. A median
across grabs is therefore not a mass balance.

The mass-domain question — how much elemental P did coprecipitation
remove from the whole event, and is that plausible against the P
actually injected — needs a discharge-weighted integral over the sampled
curve, the same construction §6 already uses to get discharge from the
NaCl mass balance (`hydro_probe`'s `integral_Ct_mgLs`, `pracma::trapz`):

$$ m_{P,\mathrm{cop}} = Q \int_0^{T} C_{P,\mathrm{cop}}(t)\,dt
   \qquad
   m_{\mathrm{SRP}} = Q \int_0^{T} C_{\mathrm{corr}}(t)\,dt $$

with $Q$ the adopted event discharge (`discharge_Ls`, treated as steady
over the duration of a single slug event, matching §6's own convention)
and the integral taken by trapezoidal rule over the sampled
`time_since_release_s`. Two ratios follow:

$$ f_{\mathrm{mass}} = \frac{m_{P,\mathrm{cop}}}{m_{\mathrm{SRP}}}
   \qquad
   \%_{\mathrm{injectate}} = 100\,\frac{m_{P,\mathrm{cop}}}{m_{P,\mathrm{added}}} $$

`f_mass` is the mass-weighted analogue of `frac_of_srp` and is the
number to compare it against. `%_injectate` is an independent physical
check that does not use `conc_corr_ugL` at all — it asks what fraction
of the elemental P mass released at the addition point was removed by
coprecipitation before it passed the sampling station. `m_P,added`
converts the phosphate-salt mass on the field sheet (`added_mass_PO4_g`,
`molar_mass_P_nutrient`, `nutrient_addition_phosphate_data.csv`) to
elemental P using the same molar-mass ratio, not the salt mass itself.

`p_coprec_ugL` is integrated as measured, not clamped at zero: a
negative value is a real model output at `flag_delta_ca_neg` events (the
reach gained calcium — §9.5 already documents the model is not
parameterised for that direction), and clipping it would silently
overstate the correction there. `conc_corr_ugL` is clamped at zero in
its integral, matching the convention already used for `nacl_mgL_grab`
in §6 (a background-corrected concentration can dip slightly negative
from measurement noise around a zero baseline; a transported mass cannot
be physically negative).

```{r coprec-mass-balance}
p_added <- p_raw %>%
  distinct(stream, date, added_mass_PO4_g, molar_mass_P_nutrient)
stopifnot(!anyDuplicated(p_added[c("stream", "date")]))

coprec_mass_balance <- btc_nutrients %>%
  filter(solute == "SRP", !is.na(p_coprec_ugL)) %>%
  select(event_id, stream, date, time_since_release_s, conc_corr_ugL,
         p_coprec_ugL, flag_delta_ca_neg) %>%
  left_join(events %>% select(event_id, discharge_Ls, flag_discharge_invalid),
            by = "event_id") %>%
  left_join(p_added, by = c("stream", "date")) %>%
  arrange(event_id, time_since_release_s) %>%
  group_by(event_id) %>%
  summarise(
    n                      = n(),
    t_span_s               = diff(range(time_since_release_s, na.rm = TRUE)),
    discharge_Ls           = first(discharge_Ls),
    flag_discharge_invalid = first(flag_discharge_invalid),
    flag_delta_ca_neg      = first(flag_delta_ca_neg),
    # conc_corr_ugL can be NA on a grab even when p_coprec_ugL is
    # defined (a bad concentration reading does not stop the
    # geochemical calc, which does not use conc_corr_ugL at all).
    # A single NA in the window propagates through pmax()/trapz()
    # to NA for mass_srp_exported_ug and therefore frac_mass_coprec,
    # while mass_p_coprec_ug and pct_of_injectate_coprec (below) are
    # unaffected -- this flag names that asymmetry instead of
    # leaving it to be reverse-engineered from a bare NA.
    flag_conc_corr_na_in_window = anyNA(conc_corr_ugL),
    added_mass_PO4_g       = first(added_mass_PO4_g),
    molar_mass_P_nutrient  = first(molar_mass_P_nutrient),
    # discharge-weighted trapezoidal integral over the sampled curve; NA
    # where there are too few points to integrate or no valid discharge
    # (mirrors the guard already used for integral_Ct_mgLs in Sec. 6).
    mass_srp_exported_ug = if (n >= 3 && t_span_s > 0 && !is.na(discharge_Ls))
                             discharge_Ls *
                               pracma::trapz(time_since_release_s, pmax(conc_corr_ugL, 0))
                           else NA_real_,
    mass_p_coprec_ug     = if (n >= 3 && t_span_s > 0 && !is.na(discharge_Ls))
                             discharge_Ls *
                               pracma::trapz(time_since_release_s, p_coprec_ugL)
                           else NA_real_,
    .groups = "drop"
  ) %>%
  mutate(
    frac_mass_coprec        = mass_p_coprec_ug / mass_srp_exported_ug,
    mass_P_added_g          = added_mass_PO4_g * (M_P / molar_mass_P_nutrient),
    pct_of_injectate_coprec = 100 * (mass_p_coprec_ug / 1e6) / mass_P_added_g,
    flag_insufficient_samples     = n < 3,
    flag_coprec_exceeds_injectate = !is.na(pct_of_injectate_coprec) &
                                     pct_of_injectate_coprec > 100
  ) %>%
  arrange(desc(frac_mass_coprec))

coprec_mass_balance %>%
  select(event_id, n, discharge_Ls, frac_mass_coprec, pct_of_injectate_coprec,
         flag_delta_ca_neg, flag_discharge_invalid, flag_insufficient_samples,
         flag_coprec_exceeds_injectate, flag_conc_corr_na_in_window) %>%
  print(n = Inf)

write_csv(coprec_mass_balance, here("data_derived", "coprec_mass_balance.csv"))

dict_coprec_mass_balance <- tribble(
  ~column, ~unit, ~description,
  "event_id",                     "-",     "Event key",
  "n",                             "-",     "Number of SRP grabs with a defined p_coprec_ugL",
  "t_span_s",                      "s",     "Time span covered by those grabs; integration domain",
  "discharge_Ls",                  "L/s",   "Adopted event discharge (events$discharge_Ls), treated as steady over the event",
  "mass_srp_exported_ug",          "ug",    "discharge_Ls * trapz(time_since_release_s, pmax(conc_corr_ugL, 0)); background-corrected SRP mass transported past the station over the sampled window",
  "mass_p_coprec_ug",              "ug",    "discharge_Ls * trapz(time_since_release_s, p_coprec_ugL); P mass removed by coprecipitation over the sampled window. Not clamped at zero -- see prose",
  "frac_mass_coprec",              "-",     "mass_p_coprec_ug / mass_srp_exported_ug; mass-weighted, event-integrated analogue of the per-grab median frac_of_srp in Sec. 9.5",
  "added_mass_PO4_g",              "g",     "Mass of the phosphate salt added, nutrient_addition_phosphate_data.csv",
  "molar_mass_P_nutrient",         "g/mol", "Molar mass of the salt added (type_P)",
  "mass_P_added_g",                "g",     "added_mass_PO4_g * (M_P / molar_mass_P_nutrient); elemental P mass injected",
  "pct_of_injectate_coprec",       "%",     "100 * (mass_p_coprec_ug/1e6) / mass_P_added_g; independent check, does not use conc_corr_ugL",
  "flag_delta_ca_neg",             "-",     "Carried from Sec. 9.5: TRUE when the reach gained calcium (model not parameterised for that direction)",
  "flag_discharge_invalid",        "-",     "Carried from Sec. 6: TRUE when no valid discharge estimate exists for this event",
  "flag_insufficient_samples",     "-",     "TRUE when n < 3 (cannot integrate); mass columns are NA",
  "flag_coprec_exceeds_injectate", "-",     "TRUE when pct_of_injectate_coprec > 100 -- physically impossible, flags a remaining error rather than a real effect",
  "flag_conc_corr_na_in_window",   "-",     "TRUE when any conc_corr_ugL in the window is NA; frac_mass_coprec is then NA but pct_of_injectate_coprec (independent of conc_corr_ugL) is not -- expected, not a bug"
)

write_csv(dict_coprec_mass_balance,
          here("data_derived", "dictionary", "dict_coprec_mass_balance.csv"))

dict_coprec_mass_balance
```

> **QAQC note.** `frac_mass_coprec` is the number that should be
> compared to the per-grab median in §9.5, not treated as a replacement
> for it — a large gap between the two says the loss is concentrated in
> part of the curve the median under- or over-weights, which is itself
> worth reporting. `pct_of_injectate_coprec` is the stronger check: it
> is built from the addition record alone (`added_mass_PO4_g`,
> `molar_mass_P_nutrient`) and the coprecipitation model, with no
> dependence on the measured downstream SRP series, so it cannot inherit
> an error from `conc_corr_ugL`. Any
> `flag_coprec_exceeds_injectate == TRUE` means the model is predicting
> more P removed than was released — physically impossible — and should
> be read as evidence of a remaining unit or join error, not as a real
> effect, per the same standard applied to `SIGMA_P` and the PHREEQC
> unit fixes earlier in this section.
>
> `flag_conc_corr_na_in_window` explains the one asymmetric case:
> `frac_mass_coprec` can be `NA` while `pct_of_injectate_coprec` is not,
> because a single `NA` in `conc_corr_ugL` propagates through the
> `mass_srp_exported_ug` integral but `mass_p_coprec_ug` never uses
> `conc_corr_ugL` at all. This is expected behaviour, not a second bug.
>
> **A note on circularity.** `p_coprec_ugL` — and therefore
> `conc_corr_coprec_ugL` — is a deterministic function of `delta_ca_mgL`
> (§9.2). Any uptake metric computed from the corrected series (Sw, Vf,
> areal uptake, or otherwise) therefore carries a component that is
> mechanically linear in ΔCa by construction, independent of any real
> biological or abiotic process. Comparing that metric, once corrected,
> back against ΔCa or calcium transfer and reporting agreement as
> confirmation of the coprecipitation mechanism is circular — the
> correction was built from ΔCa, so some relationship between the two is
> guaranteed rather than discovered. The evidence that is not circular
> is the relationship measured on the *uncorrected* series (nothing in
> `conc_corr_ugL` depends on ΔCa) and the independent mass-balance check
> above (`pct_of_injectate_coprec`, built from the addition record, not
> from ΔCa or the corrected concentration). A relationship that is
> absent uncorrected and appears only after applying the correction is
> particularly worth treating with suspicion for this reason before
> being reported as a finding — it is at least as consistent with the
> correction's own linear structure as with a newly-revealed process.

# 10. Consolidated master dataset for inverse TSM

Earlier versions of this section built one dataset for the RA reach
only, and explicitly excluded 2023-10-05 (both stations) from it: that
day's SRP series is erratic (swings from -3 to +120 µg/L point to point,
no coherent rise-fall shape), a broken NaCl-nutrient pairing traced to
the same pre-tracer instrument step documented in Section 3, and not to
co-precipitation (`03` §1.1 shows the predicted correction is the same
negligible order for both RA dates). That exclusion contradicted this
project's own standing rule: quantify and flag uncertainty, never
silently drop a stream. **It is reversed here.**

This section now builds one master table across **every** event with a
nutrient series — nothing is dropped. Discharge and reach geometry come
from each event's own individually-adopted value in `events` (Section
6), \*\*with no override, including for RA_20230906_N/\_P\*\*: each
slug's own hydraulics (`water_velocity_ms`, `stream_depth_m`, and
whatever an inverse model derives from its own BTC) must stay internally
consistent with the same discharge that was used to derive them in
Section 6. As of Section 6's latest correction, `discharge_Ls` for
RA_20230906_N, RA_20230906_P, and RA_20231005_downstream is now the
probe (handheld trapezoidal) estimate rather than the manual Excel value
— each still carries its own individually-adopted number, not a shared
one. What the two RA_20230906 slugs' reach genuinely shares that day —
one physical river, one set of survey geometry — is a separate,
explicitly-labelled reference pair, `discharge_Ls_day_mean`/`_sd` (mean
and sd of the two slugs' own adopted `discharge_Ls`), populated only for
those two rows and not substituted into `discharge_Ls` anywhere. To make
the logger-vs-probe-vs-manual picture fully inspectable from this table
alone — not just from `events.csv` — the raw per-source estimates
(`discharge_Ls_logger`, `discharge_Ls_probe`, `discharge_Ls_manual`),
the adopted method, and the reason the logger was rejected where it was
are carried forward here too. Every known data-quality problem —
RA_20231005 upstream's incomplete mixing, RA_20231005's broken pairing
(both stations), missing baselines, a peak barely above background,
sparse recessions, discharge that could not be estimated at all — is
carried forward as an explicit flag column instead. Whether to use,
weight, or exclude a flagged event is a modelling decision made
downstream of this table, not a decision this pipeline makes on anyone's
behalf; RA_20231005_upstream in particular is expected to sit unused as
a reference row for now.

```{r ra-reach-discharge}
# Day-level reference discharge for the two RA 2023-09-06 slugs, ~2h15min
# apart, same physical reach. Built from each slug's own ADOPTED
# discharge_Ls in `events` -- not a hardcoded pair of numbers -- so this
# stays consistent automatically if Section 6's source decision for either
# slug ever changes again (it already has once: manual -> probe). This
# mean is NOT substituted into either event's own discharge_Ls below --
# each slug's own velocity/depth were derived from its own value, and
# overwriting it would break that internal consistency. It is instead
# kept as a day-level REFERENCE for what the river's discharge looked
# like that day (mean of two independent estimates of the same
# underlying flow, spread carried as uncertainty rather than picking
# one) -- available if a future step wants a single characterising Q
# for 2023-09-06, not used by this table itself.
ra_q_lookup <- events %>%
  filter(event_id %in% c("RA_20230906_N", "RA_20230906_P")) %>%
  select(event_id, discharge_Ls)

stopifnot(nrow(ra_q_lookup) == 2, !anyNA(ra_q_lookup$discharge_Ls))

RA_Q_VALUES   <- set_names(ra_q_lookup$discharge_Ls, ra_q_lookup$event_id)
RA_Q_MEAN_LS  <- mean(RA_Q_VALUES)
RA_Q_SD_LS    <- sd(RA_Q_VALUES)

tibble(discharge_Ls_mean = RA_Q_MEAN_LS, discharge_Ls_sd = RA_Q_SD_LS,
       cv_pct = 100 * RA_Q_SD_LS / RA_Q_MEAN_LS)
```

> **QAQC note.** `sd()` on two points is `abs(diff(x))/sqrt(2)` —
> reported here because it is the standard, honest way to carry a
> 2-value spread forward, not because n=2 supports a distributional
> claim. The mean/sd printed above is carried in the master table below
> as `discharge_Ls_day_mean`/`_sd`, a reference value only;
> `discharge_Ls` for RA_20230906_N and RA_20230906_P stays each slug's
> own adopted value, exactly as in `events`/`hydraulics`.

```{r master-tsm-export}
# Day-level reference discharge for RA 2023-09-06 only (see prose and
# QAQC note above) -- NA for every other event, including RA_20231005.
# Not used to compute discharge_Ls, water_velocity_ms, or stream_depth_m
# anywhere in this table.
ra_day_reference <- tibble(
  event_id                  = names(RA_Q_VALUES),
  discharge_Ls_day_mean     = RA_Q_MEAN_LS,
  discharge_Ls_day_mean_sd  = RA_Q_SD_LS
)

master_hydraulics <- events %>%
  select(event_id, discharge_Ls, reach_length_m, reach_mean_width_m,
         discharge_source, discharge_method, reason_discharge,
         discharge_Ls_logger, discharge_Ls_probe, discharge_Ls_manual,
         flag_discharge_invalid, flag_mixing_incomplete,
         flag_no_baseline, flag_weak_signal, flag_sparse_recession) %>%
  left_join(ra_day_reference, by = "event_id") %>%
  mutate(
    # RA 2023-10-05, both stations: erratic SRP series, broken NaCl
    # pairing (see prose above). NOT excluded -- flagged.
    # RA_20231005_downstream already carries flag_no_baseline == TRUE
    # (Section 6, contaminated start anchor) and RA_20231005_upstream
    # already carries flag_mixing_incomplete == TRUE; this flag is the
    # additional, independent problem with the nutrient pairing itself.
    flag_broken_pairing = event_id %in% c("RA_20231005_upstream",
                                          "RA_20231005_downstream")
  )

# guard: one hydraulics row per event, so the join below cannot
# multiply rows.
stopifnot(!anyDuplicated(master_hydraulics$event_id))

master_tsm <- btc_nutrients %>%
  left_join(master_hydraulics, by = "event_id") %>%
  arrange(event_id, solute, time_since_release_s)

# guard: every event in btc_nutrients survives the join (left_join
# cannot drop left-table rows, but a name/type mismatch could still
# silently produce all-NA hydraulics -- catch that here), and wherever
# discharge_Ls itself is NA, flag_discharge_invalid says so.
stopifnot(setequal(unique(master_tsm$event_id), unique(btc_nutrients$event_id)))
stopifnot(all(master_tsm$flag_discharge_invalid[is.na(master_tsm$discharge_Ls)]))

write_csv(master_tsm, here("data_derived", "master_tsm.csv"))
saveRDS(master_tsm,   here("data_derived", "master_tsm.rds"))

master_tsm %>%
  distinct(event_id, discharge_Ls, discharge_source, discharge_method,
           discharge_Ls_logger, discharge_Ls_probe, discharge_Ls_manual,
           discharge_Ls_day_mean, flag_discharge_invalid,
           flag_mixing_incomplete, flag_no_baseline, flag_weak_signal,
           flag_sparse_recession, flag_broken_pairing) %>%
  arrange(event_id) %>%
  print(n = Inf)

dict_master_tsm_extra <- tribble(
  ~column, ~unit, ~description,
  "discharge_Ls",             "L/s", "Adopted event discharge, taken straight from events$discharge_Ls with no override for any event -- source given by discharge_source/discharge_method",
  "reach_length_m",            "m",   "events$reach_length_m, unchanged for every event",
  "reach_mean_width_m",        "m",   "events$reach_mean_width_m, unchanged for every event",
  "discharge_source",          "-",   "logger / probe / manual_excel / NA, from events, unchanged for every event",
  "discharge_method",          "-",   "Method behind the adopted discharge_Ls (logger_trapz_truepeak / probe_trapz_endrising / manual_fixed_background)",
  "reason_discharge",          "-",   "Why the logger was not used, where it was not; NA when the logger is the adopted source",
  "discharge_Ls_logger",       "L/s", "Raw logger-based trapezoidal estimate (Sec. 6), independent of which source was adopted; NA where no usable logger series exists",
  "discharge_Ls_probe",        "L/s", "Raw handheld/probe-based trapezoidal estimate (Sec. 6), independent of which source was adopted; currently the adopted source for RA_20230906_N/_P and RA_20231005_downstream",
  "discharge_Ls_manual",       "L/s", "Manual Excel value (fixed background, negatives retained), Sec. 6; kept as reference only -- not the adopted source for any event as of this revision",
  "discharge_Ls_day_mean",     "L/s", "mean(each slug's own adopted discharge_Ls), RA_20230906_N/_P only; NA for every other event. A reference value for the river's discharge that day -- NOT used to compute discharge_Ls or anything derived from it in this table",
  "discharge_Ls_day_mean_sd",  "L/s", "sd() of the same two adopted values, RA_20230906_N/_P only; NA elsewhere",
  "flag_discharge_invalid",    "-",   "TRUE when no usable discharge exists for this event -- currently RA_20231005_upstream only",
  "flag_mixing_incomplete",    "-",   "TRUE for RA_20231005_upstream: logger sampled the plume core, not a well-mixed cross-section, so the dilution-gauging premise fails",
  "flag_no_baseline",          "-",   "TRUE when the window carries no clean pre-release baseline (RA_20230906_N, RA_20231005_downstream)",
  "flag_weak_signal",          "-",   "TRUE when the adopted trapezoidal estimate's peak is only marginally above background noise (RA_20230906_P: ~3%)",
  "flag_sparse_recession",     "-",   "TRUE when one unsampled interval holds >30% of the handheld trapezoidal area",
  "flag_broken_pairing",       "-",   "TRUE for RA_20231005 (both stations): SRP series is erratic (-3 to +120 ug/L point to point, no coherent shape) and the NaCl pairing is at or near zero for nearly the whole grab window, per Section 3 and 03 Sec 1.1. NOT excluded from this table -- using, weighting, or dropping these rows is a modelling decision, not a pipeline decision"
)
write_csv(bind_rows(dict_btc_nutrients, dict_coprec, dict_master_tsm_extra),
          here("data_derived", "dictionary", "dict_master_tsm.csv"))

nrow(master_tsm)
dict_master_tsm_extra
```

> **QAQC note.** This table replaces the RA-only `ra_reach_tsm` /
> `RA_reach_tsm.csv` from earlier versions of this section, which
> excluded RA_20231005 outright. Every event is here now, and
> `flag_broken_pairing`, `flag_discharge_invalid`,
> `flag_mixing_incomplete`, `flag_no_baseline`, and `flag_weak_signal`
> are how a modelling step should decide what to do with a given event,
> not something this pipeline decides on its behalf.
> RA_20231005_upstream in particular is expected to sit unused, as a
> reference row, until there is a specific reason to touch it.
> `discharge_Ls_logger`/`_probe`/`_manual` are all carried here (not
> just `discharge_Ls`) specifically so this table alone supports a
> source-vs-source comparison — see the 1:1 plot in Section 6 for the
> logger-vs-probe picture across all events. Running the inverse model
> on all events, dropping the flagged ones, or running a sensitivity
> comparison with/without them is a choice for whichever conversation
> builds that model next (see the handoff).

```{r session-info}
sessionInfo()
```
