---
title: "03 — Breakthrough curve review and hysteresis"
subtitle: "Visual QAQC of the final tables, and event-scale hysteresis"
author: "Kauan Fonseca"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7
)

library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(plotly)
library(here)
```

# Scope

This document does **no processing**: it reads the three derived tables
written by `02_integration.Rmd` and renders them. If a curve looks wrong
here, the fix belongs in `02`, never in `03`. It is the final visual
sanity check before the tables are used for inverse modelling, and it
reports the nutrient-tracer hysteresis, which is a result rather than a
QAQC step.

```{r read-tables}
btc_conservative <- read_rds(here("data_derived", "btc_conservative.rds"))
btc_nutrients    <- read_rds(here("data_derived", "btc_nutrients.rds"))
events           <- read_rds(here("data_derived", "events.rds"))

nrow(btc_conservative); nrow(btc_nutrients); nrow(events)
```

---

# 1. The four curves per event

For every event this overlays, against time since release: the logger
SpC, the logger-derived NaCl, and the two nutrients (NH4-N, SRP). SpC and
NaCl are the high-resolution logger series; the nutrients are grab
samples. Each panel is one event; the y-axes are free because the four
quantities span different ranges.

```{r long-for-plot}
cons_long <- btc_conservative %>%
  transmute(event_id, time_min = time_since_release_s / 60,
            SpC = spc_uscm, NaCl = nacl_mgL) %>%
  pivot_longer(c(SpC, NaCl), names_to = "series", values_to = "value") %>%
  mutate(kind = "logger", variant = "background-corrected")

# Two variants of the same series: the background-corrected concentration
# (all solutes) and, for SRP only, the same value plus the House (1990)
# co-precipitation correction (Section 9.5/9.8 of `02`). Both are kept in
# the SRP facet, distinguished by colour, so the size of the correction is
# visible directly on the breakthrough curve rather than only in the
# parity plot of Section 1.1.
nut_long <- bind_rows(
  btc_nutrients %>%
    transmute(event_id, time_min = time_since_release_s / 60,
              series = solute, value = conc_corr_ugL, kind = "grab",
              variant = "background-corrected"),
  btc_nutrients %>%
    filter(solute == "SRP", !is.na(conc_corr_coprec_ugL)) %>%
    transmute(event_id, time_min = time_since_release_s / 60,
              series = solute, value = conc_corr_coprec_ugL, kind = "grab",
              variant = "+ coprecipitation")
) %>%
  filter(!is.na(value))

curves <- bind_rows(cons_long, nut_long) %>%
  mutate(series = factor(series, levels = c("SpC", "NaCl", "NH4-N", "SRP")))
```

```{r plot-per-event, fig.height=4}
# one interactive figure per event; each variable on its own facet row so
# the different units stay readable
make_event_plot <- function(ev) {
  d        <- curves %>% filter(event_id == ev)
  d_logger <- d %>% filter(kind == "logger")
  d_grab   <- d %>% filter(kind == "grab")

  # geom_vline(xintercept = 0) is an infinite-extent geom with no data rows
  # of its own; ggplotly's conversion does not carry it across (silently
  # drops it), so it is replaced with a geom_segment anchored to each
  # facet's own y-range (scales are free_y, so a single -Inf/Inf pair would
  # not clip correctly per panel).
  vline_df <- d %>%
    group_by(series) %>%
    summarise(y_min = min(value, na.rm = TRUE),
              y_max = max(value, na.rm = TRUE), .groups = "drop")

  # background reference per panel: the pre-release/pre-addition level each
  # series is judged against. SpC's is the handheld-probe reading at release
  # (background_spc_probe, 01); NH4-N/SRP are the pre-addition grab
  # background computed in 02 (nh4_background_ugL / srp_background_ugL).
  # NaCl has none plotted — it is already background-corrected in 02, so its
  # own baseline is 0 by construction, not a separate reference value.
  ev_row <- events %>% filter(event_id == ev)
  bg_df <- tibble(
    series     = factor(c("SpC", "NH4-N", "SRP"), levels = levels(d$series)),
    background = c(ev_row$background_spc_probe, ev_row$nh4_background_ugL,
                   ev_row$srp_background_ugL)
  ) %>%
    filter(!is.na(background))

  p <- ggplot(d, aes(time_min, value)) +
    geom_segment(data = vline_df,
                 aes(x = 0, xend = 0, y = y_min, yend = y_max),
                 inherit.aes = FALSE, linetype = "dashed", colour = "darkgreen")

  if (nrow(bg_df) > 0) {
    p <- p + geom_hline(data = bg_df, aes(yintercept = background),
                        inherit.aes = FALSE, linetype = "dotted",
                        colour = "steelblue", linewidth = 0.4)
  }

  # the logger (SpC/NaCl) layer is added only when this event actually has
  # logger rows: SR has no logger at all (see readme.docx field note), so
  # d_logger is zero-row for SR_20231011_single, and geom_line() on an
  # empty data frame crashes plotly's to_basic.GeomLine() during ggplotly()
  # conversion (Error in order(): argument 1 is not a vector).
  if (nrow(d_logger) > 0) {
    p <- p + geom_line(data = d_logger, linewidth = 0.3)
  }

  # grabs are sparse (NH4-N/SRP), so a connecting line makes the shape of
  # the curve readable between points, not just the isolated values. Guarded
  # the same way as the logger layer above: events with has_nutrients ==
  # FALSE (e.g. SA_20231010, RA_20231005_upstream) have zero grab rows, and
  # geom_line()/geom_point() on empty data crashes ggplotly() conversion.
  if (nrow(d_grab) > 0) {
    p <- p +
      geom_line(data = d_grab, aes(colour = variant), linewidth = 0.3) +
      geom_point(data = d_grab, aes(colour = variant), size = 1) +
      scale_colour_manual(values = c("background-corrected" = "firebrick",
                                     "+ coprecipitation"     = "darkorange"),
                          name = NULL)
  }

  p <- p +
    facet_wrap(~ series, scales = "free_y", ncol = 1, strip.position = "left") +
    labs(x = "Time since release (min)", y = NULL, title = ev,
         caption = "Dotted blue = pre-release/pre-addition background") +
    theme_minimal(base_size = 9) +
    theme(strip.placement = "outside", legend.position = "bottom")

  ggplotly(p, height = 520)
}

event_ids <- events %>% filter(has_nutrients | has_logger) %>% pull(event_id) %>% sort()
```

```{r render-event-plots, results='asis'}
# htmltools stitches the per-event widgets into the document
htmltools::tagList(lapply(event_ids, make_event_plot))
```

## 1.1 Corrected vs uncorrected SRP

`conc_corr_ugL` is SRP after background subtraction only.
`conc_corr_coprec_ugL` adds back the House (1990) co-precipitated
fraction (`p_coprec_ugL`), reconstructing what SRP would have been
without abiotic removal to CaCO₃. If the two series diverge visibly, the
correction is doing real work; if they overlay, it is not — either
because too little calcite is forming to matter for that event, or
because the correction itself is negligible at these activities.

```{r srp-corrected-parity}
srp_pts <- btc_nutrients %>%
  filter(solute == "SRP", !is.na(conc_corr_ugL), !is.na(conc_corr_coprec_ugL))

ggplotly(
  ggplot(srp_pts, aes(conc_corr_ugL, conc_corr_coprec_ugL, colour = event_id)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(size = 1.2, alpha = 0.7) +
    labs(x = "SRP, background-corrected (µg/L)",
         y = "SRP, + co-precipitation correction (µg/L)",
         title = "Every point on the 1:1 line means the coprec correction changed nothing there") +
    theme_minimal(base_size = 9)
)
```

```{r srp-corrected-magnitude}
srp_pts %>%
  group_by(event_id) %>%
  summarise(
    n = n(),
    p_coprec_median_ugL = median(p_coprec_ugL, na.rm = TRUE),
    frac_of_srp_median  = median(p_coprec_ugL / conc_corr_ugL, na.rm = TRUE),
    delta_ca_mgL        = first(delta_ca_mgL),
    .groups = "drop"
  ) %>%
  arrange(desc(frac_of_srp_median))
```

> **QAQC note.** `frac_of_srp_median` is of order 1e-8 to 1e-10 for
> every event in this dataset — the correction moves SRP by a fraction
> of a part-per-billion of itself, not a visible amount, regardless of
> how much calcium is being lost (`delta_ca_mgL`, 0.3-0.4 mg/L for most
> events, comparable across the board). Concretely: this rules out
> co-precipitation as the explanation for why `RA_20231005_downstream`'s
> SRP series looks noisier than `RA_20230906_P`'s (Section 1) — the
> predicted correction for both is of the same tiny order, so it cannot
> be what separates a coherent curve from an erratic one. That said, a
> correction this uniformly small, for the one process this manuscript
> is built to quantify, is itself worth checking against the House
> (1986/1990) source: `a_PO4`/`a_HPO4` come from PHREEQC (`R/phreeqc_speciation.R`)
> and are extremely small on their own (~1e-14 to 1e-9), which is
> plausible for PO₄³⁻ activity at this pH and total-P, but the Langmuir
> constants `k1_langmuir`/`k2_langmuir` (from `A1`, `E1`, `A2`, `E2` in
> `coprec-constants`) were not re-derived here and are not verified
> against the source units. This is flagged, not corrected — changing
> those constants is not a call to make without the original paper in
> hand.

# 2. Nutrient vs conservative tracer, overlaid

A more compact view: nutrients and NaCl on the same normalised axis, so
the timing of the nutrient relative to the conservative tracer is visible
directly. A nutrient that peaks later or lower than NaCl is being
retained; one that tracks NaCl is behaving conservatively.

```{r overlay-normalised}
rescale01 <- function(x) {
  r <- diff(range(x, na.rm = TRUE))
  if (!is.finite(r) || r == 0) return(rep(NA_real_, length(x)))
  (x - min(x, na.rm = TRUE)) / r
}

overlay <- btc_nutrients %>%
  filter(!is.na(conc_corr_ugL), !is.na(nacl_mgL)) %>%
  group_by(event_id, solute) %>%
  mutate(nutrient_norm = rescale01(conc_corr_ugL),
         nacl_norm     = rescale01(nacl_mgL),
         time_min      = time_since_release_s / 60) %>%
  ungroup()

p_overlay <- overlay %>%
  pivot_longer(c(nutrient_norm, nacl_norm),
               names_to = "series", values_to = "value") %>%
  mutate(series = recode(series, nutrient_norm = "nutrient", nacl_norm = "NaCl")) %>%
  ggplot(aes(time_min, value, colour = series)) +
  geom_line(linewidth = 0.3) +
  geom_point(size = 0.8) +
  facet_wrap(event_id ~ solute, scales = "free_x") +
  labs(x = "Time since release (min)", y = "Normalised (0–1)",
       colour = NULL) +
  theme_minimal(base_size = 8)

ggplotly(p_overlay)
```

# 3. Hysteresis of the nutrient breakthrough curves

Following the normalised-loop approach of Lloyd et al. (2016b), adapted
to a slug-addition setting in which the conservative tracer replaces
discharge on the x-axis. Both variables are min–max normalised within
each event and the vertical separation between the rising and falling
limbs is evaluated at equally spaced tracer levels, restricted to the
range covered by both limbs. Positive values are clockwise loops
(nutrient leads the tracer); negative are anticlockwise (nutrient lags —
the signature of exchange with the transient storage zone).

```{r hysteresis-fn}
hysteresis_index <- function(df, n_levels = 9, min_overlap = 0.2) {
  rise <- df %>% filter(curve_limb == "rising")  %>% arrange(x_norm)
  fall <- df %>% filter(curve_limb == "falling") %>% arrange(x_norm)
  if (nrow(rise) < 3 || nrow(fall) < 3)
    return(tibble(level = NA_real_, hi = NA_real_, overlap = NA_real_))
  lo <- max(min(rise$x_norm), min(fall$x_norm))
  up <- min(max(rise$x_norm), max(fall$x_norm))
  if (!is.finite(lo) || !is.finite(up) || (up - lo) < min_overlap)
    return(tibble(level = NA_real_, hi = NA_real_, overlap = up - lo))
  lv <- seq(lo, up, length.out = n_levels)
  tibble(level = lv,
         hi = approx(rise$x_norm, rise$y_norm, xout = lv, ties = mean)$y -
              approx(fall$x_norm, fall$y_norm, xout = lv, ties = mean)$y,
         overlap = up - lo)
}
```

```{r hysteresis}
hyst_input <- btc_nutrients %>%
  filter(!is.na(nacl_mgL), !is.na(conc_corr_ugL),
         curve_limb %in% c("rising", "falling")) %>%
  group_by(event_id, solute) %>%
  mutate(x_norm = rescale01(nacl_mgL),
         y_norm = rescale01(conc_corr_ugL)) %>%
  ungroup()

hyst <- hyst_input %>%
  group_by(event_id, solute) %>%
  group_modify(~ hysteresis_index(.x)) %>%
  ungroup()

hyst_summary <- hyst %>%
  group_by(event_id, solute) %>%
  summarise(
    n_levels     = sum(!is.na(hi)),
    limb_overlap = first(overlap),
    HI_mean      = if (any(!is.na(hi))) mean(hi, na.rm = TRUE) else NA_real_,
    HI_max       = if (any(!is.na(hi))) hi[which.max(abs(replace_na(hi, 0)))] else NA_real_,
    .groups = "drop"
  ) %>%
  arrange(desc(abs(HI_mean)))

hyst_summary
```

```{r hysteresis-loops}
ggplotly(
  btc_nutrients %>%
    filter(!is.na(nacl_mgL), !is.na(conc_corr_ugL)) %>%
    arrange(event_id, solute, time_since_release_s) %>%
    ggplot(aes(nacl_mgL, conc_corr_ugL, colour = curve_limb)) +
    geom_path(colour = "grey70", linewidth = 0.3) +
    geom_point(size = 1.2) +
    facet_wrap(event_id ~ solute, scales = "free") +
    labs(x = "NaCl (mg/L)", y = "Background-corrected nutrient (µg/L)") +
    theme_minimal(base_size = 8)
)
```

```{r export-hysteresis}
write_csv(hyst,         here("data_derived", "hysteresis_levels.csv"))
write_csv(hyst_summary, here("data_derived", "hysteresis_summary.csv"))
```

> **Note.** `limb_overlap` is a quality field: an index over 25 % of the
> tracer range is not comparable to one over 80 %, and events below
> `min_overlap = 0.2` return NA rather than a number built on
> extrapolation. CC 2023-10-06 and RC 2023-10-07 fall out here — their
> rising limbs were sampled only near the base of the curve — and are
> reported as NA.

# 4. Discharge cross-check

```{r discharge-check}
events %>%
  filter(!is.na(discharge_ratio)) %>%
  ggplot(aes(reorder(event_id, discharge_ratio), discharge_ratio)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_col(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(x = NULL, y = "Discharge ratio (logger / probe)") +
  theme_minimal(base_size = 10)
```

```{r session-info}
sessionInfo()
```
