# 9. CaCO₃ co-precipitation correction for SRP

## 9.1 Why the correction is needed

In alkaline karst streams part of the dissolved inorganic phosphorus is
removed abiotically, adsorbed onto calcite as it precipitates, rather
than assimilated by organisms. In the breakthrough curve that removal is
indistinguishable from biological uptake: both make phosphorus disappear
faster than the conservative tracer. Left uncorrected it inflates every
phosphorus uptake metric, and the inflation is largest exactly where
calcite precipitation is strongest — a systematic bias correlated with
the variable of interest, not noise.

## 9.2 What the correction does

Following House (1986, 1990), the co-precipitated P at each grab is

$$ n_{P,\mathrm{cop}} = \Delta n_{\mathrm{Ca}}\,(\sigma\,N_A\,d)\,h $$

where $\Delta n_{\mathrm{Ca}}$ is the net loss of dissolved Ca²⁺ across
the reach, $\sigma$ the maximum surface density of P on precipitating
calcite, $d$ the molecular surface area of calcite, $N_A$ Avogadro's
number, and $h$ the fraction of dissolved phosphate liable to sorb, from
the two-component Langmuir model

$$ h = \frac{K_1 a_{\mathrm{PO}_4^{3-}} + K_2 a_{\mathrm{HPO}_4^{2-}}}
             {1 + K_1 a_{\mathrm{PO}_4^{3-}} + K_2 a_{\mathrm{HPO}_4^{2-}}} $$

with activities from PHREEQC and $K_1$, $K_2$ temperature-dependent
through Arrhenius expressions.

**Where each term comes from.** The activities are computed from the
**downstream** solution — the water passing the sampling station is what
must be speciated — using downstream pH, temperature and major ions,
with `ca_mgL_down` rather than any upstream-downstream mean.
$\Delta n_{\mathrm{Ca}}$ comes from a synoptic pair: upstream and
downstream calcium sampled on arrival at the reach, before the nutrient
release. It measures the calcite precipitation the reach was sustaining
under ambient conditions, so it is one value per event, and the
within-event variation in $n_{P,\mathrm{cop}}$ comes entirely from $h$,
which tracks phosphate speciation as the slug passes. This follows from
the design: a single ambient pair characterises the reach, and the model
scales it by how much phosphate is present at each instant.

Two consequences are stated rather than worked around. The calcium pair
predates the release, so it cannot capture any change in calcite dynamics
caused by the addition itself. And being a single pair, the correction
carries no within-event uncertainty — its uncertainty is entirely between
events.

**Data provenance note.** The `ph` and `temp_c` columns of the nutrient
spreadsheets originally carried the upstream reading for some events, a
field-sheet transcription error corrected before the speciation was run.
Since 0.1 pH unit and 1 °C both shift $K_1$, $K_2$ and the
PO₄³⁻/HPO₄²⁻ partition measurably, the guard in
`R/phreeqc_speciation.R` re-checks every event against the synoptic
upstream table and warns if any still matches the upstream value exactly.

## 9.3 How it enters the pipeline, and why not as an overwrite

Under TASCC the correction feeds a per-sample regression of
$S_{w\text{-add}}$ against the geometric mean concentration, so each grab
yields its own uptake length and a corrected point simply replaces an
uncorrected one. Under inverse transient-storage modelling the corrected
series **is** the observation vector: the whole curve is fitted at once,
and a correction applied unevenly along it changes the shape the model is
matching, not just the value at a point. Overwriting would make the two
analyses non-comparable and irreversible.

The correction is therefore added as parallel columns and
`conc_corr_ugL` is never modified, so both series can be run through the
inverse model and the difference attributed to the abiotic pathway.

**Naming.** The suffix `_corr` already means "background-corrected"
throughout this pipeline (`conc_corr_ugL`, `spc_corr_uscm`,
`spc_corr_uscm_grab`). The co-precipitation column extends that same
series rather than introducing a second meaning, so it is
`conc_corr_coprec_ugL`: the background-corrected concentration, plus the
co-precipitated fraction. It is `NA` for NH4-N, like every other
SRP-specific column in this long-format table.

## 9.4 Geochemical check

Evidence that calcite was actively precipitating should not come from the
same model that assumes it. Calcite saturation indices are not used: the
speciation input carries no carbonate species, and the only alkalinity
available is a campaign mean rather than an event-specific measurement,
so a computed index would rest on a value that was not measured on the
day.

Two independent lines are used instead, and they answer different
questions.

**Calcium is spatial.** Both samples were taken on arrival at the reach,
one at each station, so `ca_response` is a genuine synoptic comparison
and measures net loss of dissolved calcium along the reach — the quantity
the House model requires. This is the line that establishes whether the
mechanism operated.

**pH, dissolved oxygen and specific conductance are temporal.** The
upstream reading is from arrival and the downstream reading from the end
of sampling, so these compare the start and end of the experiment rather
than two stations. Rising pH together with rising DO is the signature of
autotrophic activity, which draws down CO₂ and drives the supersaturation
that makes calcite precipitate. Falling specific conductance is
consistent with removal of Ca²⁺ and HCO₃⁻, and is the same diel drift the
drifting baseline of Section 3 corrects for.

The distinction matters: a temporal change at one station does not by
itself demonstrate loss along the reach. Calcium demonstrates that; the
temporal indicators say whether the system was behaving as an active
autotrophic-calcite system while it happened. The correction is applied
where calcium shows net loss and the temporal indicators are consistent
with that mechanism.

```{r coprec-geochem-check}
synoptic_up <- read_csv(here("data", "nutrients", "synoptic_upstream.csv"),
                        show_col_types = FALSE) %>%
  mutate(date = ymd(as.character(date)))

geochem_check <- p_raw %>%
  group_by(stream, date) %>%
  summarise(across(any_of(c("ph", "temp_c", "do_mgL", "ca_mgL_up",
                            "ca_mgL_down", "SpC_uScm_bck")),
                   ~ first(na.omit(.x))), .groups = "drop") %>%
  rename(ph_down = ph, temp_c_down = temp_c, spc_uscm_down = SpC_uScm_bck) %>%
  {  # DO downstream is optional: drop the term if the column is absent
    if ("do_mgL" %in% names(.)) rename(., do_mgL_down = do_mgL) else .
  } %>%
  left_join(synoptic_up, by = c("stream", "date")) %>%
  mutate(
    # ---- SPATIAL (synoptic): both Ca samples taken on arrival ----
    ca_response = ca_mgL_down / ca_mgL_up,
    ca_loss_pct = 100 * (1 - ca_response),

    # ---- TEMPORAL: upstream on arrival vs downstream at end of sampling ----
    delta_ph_expt  = ph_down - ph_up,
    delta_spc_expt = spc_uscm_down - spc_uscm_up,
    spc_change_pct = 100 * (spc_uscm_down / spc_uscm_up - 1),
    delta_do_expt  = if ("do_mgL_down" %in% names(cur_data_all()))
                       do_mgL_down - do_mgL_up else NA_real_,

    # autotrophic signature: pH rising, and DO rising where it was measured
    autotrophic_signal = delta_ph_expt > 0 &
                         (is.na(delta_do_expt) | delta_do_expt > 0),

    # the synoptic sheet and the nutrient sheet must agree on upstream Ca
    ca_up_agrees = abs(ca_mgL_up - ca_mgL_up_synoptic) < 0.5
  )

geochem_check %>%
  select(stream, date, ca_mgL_up, ca_mgL_down, ca_loss_pct,
         delta_ph_expt, delta_do_expt, spc_change_pct,
         autotrophic_signal, ca_up_agrees) %>%
  arrange(desc(ca_loss_pct)) %>%
  print(n = Inf)
```

> **QAQC note.** `ca_loss_pct` above zero means the reach lost dissolved
> calcium between the stations, which is what precipitation produces;
> below zero means it gained calcium, which precipitation cannot explain
> and dissolution can. Because calcium rests on a single sample at each
> station, a loss of a few percent sits within the analytical error of
> the determination and should not be read as signal.
>
> `autotrophic_signal` is context, not proof: it says the reach was
> photosynthesising between the two readings, which is the process that
> drives calcite supersaturation. It cannot substitute for the calcium
> response, because it compares two moments rather than two stations.
> `ca_up_agrees` guards against the synoptic sheet and the nutrient sheet
> disagreeing on the same upstream number.
>
> The synoptic pair exists only for the seven events that received
> nutrients. CB 2023-09-07, RA 2023-09-06 and SA 2023-10-10 have no
> upstream calcium and therefore receive no co-precipitation correction;
> they are not part of the phosphorus comparison.

## 9.5 Application

```{r coprec-constants}
# House (1986, 1990). sigma is the maximum surface density of P associated
# with calcite precipitation in alkaline streams; d the molecular surface area
# of calcite; A1/A2 and E1/E2 the Arrhenius coefficients of the two-component
# Langmuir constants.
#
# SIGMA_P: House (1990) reports several sigma estimates, and they cluster
# by solution type, not scatter randomly -- field/river waters run
# 0.055-0.095 umol/m2 (River Frome, River Windrush, Waterston chalk
# aquifer), while idealised synthetic Ca(HCO3)2 solutions run higher,
# 0.13-0.22 umol/m2. This is a karst STREAM, not a synthetic solution, so
# the field cluster is the physically appropriate class. 5.5e-8 mol/m2 is
# the River Frome value (House et al. 1986a; the "river water" case House
# 1990 itself highlights), and is the same value and the same reasoning
# Cohen et al. (2013, Ecological Monographs 83:2) adopted for the
# Ichetucknee River -- a spring-fed karst/calcite stream directly
# analogous to this system -- explicitly preferring the field value over
# laboratory values because natural water carries competing solutes
# (Mg, Fe, DOM) that reduce the effective sorption area, and because their
# own sensitivity check found the lab-range value inflates the correction
# ~47x and produces implausible C:P ratios. See Section 9.8 for the
# corresponding sensitivity range in this pipeline.
SIGMA_P   <- 5.5e-8        # mol/m2 -- River Frome (House 1990 / Cohen et al. 2013)
D_CALCITE <- 2.01e-19      # m2/molecule
N_AVOG    <- 6.02214076e23 # molecules/mol
R_GAS     <- 8.314         # J/mol/K
A1 <- 0.6915  ; E1 <- 18.2e3   # J/mol
A2 <- 4.361e-9; E2 <- 42.6e3   # J/mol

M_P  <- 30.973762   # g/mol
M_CA <- 40.078      # g/mol

h_langmuir2 <- function(a_PO4, a_HPO4, K1, K2) {
  num <- K1 * a_PO4 + K2 * a_HPO4
  num / (1 + num)
}
```

```{r coprec-apply}
act_path <- here("data_derived", "phreeqc_activities.csv")

if (!file.exists(act_path)) {
  warning("phreeqc_activities.csv nao encontrado: rode R/phreeqc_speciation.R. ",
          "As colunas de coprecipitacao ficarao NA.")
  coprec <- tibble(stream = character(), date = as.Date(character()),
                   sample_btc = numeric())
} else {
  coprec <- read_csv(act_path, show_col_types = FALSE) %>%
    mutate(date = as.Date(date)) %>%
    left_join(
      p_raw %>% select(stream, date, sample_btc, temp_c, ca_mgL_up, ca_mgL_down),
      by = c("stream", "date", "sample_btc")
    ) %>%
    mutate(
      T_K = temp_c + 273.15,
      k1_langmuir = A1 * exp(E1 / (R_GAS * T_K)),
      k2_langmuir = A2 * exp(E2 / (R_GAS * T_K)),
      # House's K1/K2 (Eq. 7-8) and the Langmuir isotherm (Eq. 9, House
      # 1990) are defined for [PO4^3-]/[HPO4^2-] in umol/dm^3 -- House
      # states this explicitly under Eq. 8. PHREEQC's -activities output
      # is on a mol/L numeric scale, six orders of magnitude smaller;
      # unconverted, it collapses h_adsorbed toward ~0 regardless of
      # saturation. See vignettes/02_integration.Rmd's coprec-apply
      # chunk for the same fix and full rationale.
      h_adsorbed  = h_langmuir2(a_PO4 * 1e6, a_HPO4 * 1e6, k1_langmuir, k2_langmuir),

      # synoptic pre-release pair; positive = calcium lost across the reach
      delta_ca_mgL  = ca_mgL_up - ca_mgL_down,
      delta_ca_molL = delta_ca_mgL / (1000 * M_CA),

      p_coprec_molL = delta_ca_molL * (SIGMA_P * N_AVOG * D_CALCITE) * h_adsorbed,
      p_coprec_ugL  = p_coprec_molL * M_P * 1e6,

      flag_delta_ca_neg = delta_ca_mgL < 0
    ) %>%
    select(stream, date, sample_btc, a_PO4, a_HPO4, h_adsorbed,
           k1_langmuir, k2_langmuir, delta_ca_mgL, p_coprec_ugL,
           flag_delta_ca_neg)
}

# Adding p_coprec to the background-corrected concentration reconstructs the
# SRP that would have been present without abiotic removal:
#   (measured + coprecipitated) - background
# Background subtraction and this addition commute, so order is immaterial.
btc_nutrients <- btc_nutrients %>%
  select(-any_of(c("a_PO4", "a_HPO4", "h_adsorbed", "k1_langmuir",
                   "k2_langmuir", "delta_ca_mgL", "p_coprec_ugL",
                   "flag_delta_ca_neg", "conc_corr_coprec_ugL"))) %>%
  left_join(coprec, by = c("stream", "date", "sample_btc")) %>%
  mutate(
    across(c(a_PO4, a_HPO4, h_adsorbed, k1_langmuir, k2_langmuir,
             delta_ca_mgL, p_coprec_ugL),
           ~ if_else(solute == "SRP", .x, NA_real_)),
    flag_delta_ca_neg = if_else(solute == "SRP", flag_delta_ca_neg, NA),
    conc_corr_coprec_ugL = if_else(solute == "SRP",
                                   conc_corr_ugL + p_coprec_ugL,
                                   NA_real_)
  )

btc_nutrients %>%
  filter(solute == "SRP") %>%
  group_by(event_id) %>%
  summarise(
    n = n(),
    delta_ca_mgL = first(delta_ca_mgL),
    h_median     = median(h_adsorbed, na.rm = TRUE),
    coprec_median_ugL = median(p_coprec_ugL, na.rm = TRUE),
    frac_of_srp  = median(p_coprec_ugL / conc_corr_ugL, na.rm = TRUE),
    delta_ca_neg = first(flag_delta_ca_neg),
    .groups = "drop"
  ) %>%
  arrange(desc(delta_ca_mgL)) %>%
  print(n = Inf)
```

> **QAQC note.** `frac_of_srp` is the size of the correction relative to
> the signal being corrected, and it decides how much any of this
> matters. Where it is small, the treatment of ΔCa < 0 cannot change a
> conclusion; where it approaches unity, the corrected series is largely
> model output rather than measurement, and the inverse fit should be
> reported for both series.
>
> Where `delta_ca_neg` is TRUE the reach gained calcium: calcite
> dissolved rather than precipitated. The House model returns a negative
> $n_{P,\mathrm{cop}}$ there, so the correction subtracts phosphorus,
> which is outside the direction the model was parameterised for. No
> precedent for this case exists in the BTC literature, so these events
> are flagged and carried forward, and robustness of the main conclusions
> is demonstrated across alternative treatments of them rather than one
> treatment being asserted as correct.

```{r coprec-export}
write_csv(btc_nutrients, here("data_derived", "btc_nutrients.csv"))
saveRDS(btc_nutrients,   here("data_derived", "btc_nutrients.rds"))

write_csv(geochem_check, here("data_derived", "geochem_check.csv"))

dict_coprec <- tribble(
  ~column, ~unit, ~description,
  "a_PO4",               "-",     "PO4(3-) activity, PHREEQC, downstream solution",
  "a_HPO4",              "-",     "HPO4(2-) activity, PHREEQC, downstream solution",
  "k1_langmuir",         "-",     "Two-component Langmuir constant for PO4(3-), Arrhenius in temp_c",
  "k2_langmuir",         "-",     "Two-component Langmuir constant for HPO4(2-), Arrhenius in temp_c",
  "h_adsorbed",          "-",     "Fraction of dissolved phosphate liable to sorb onto new calcite",
  "delta_ca_mgL",        "mg/L",  "ca_mgL_up - ca_mgL_down, synoptic pair on arrival; positive = Ca lost",
  "p_coprec_ugL",        "ug/L",  "Inorganic P co-precipitated with calcite, House (1990)",
  "conc_corr_coprec_ugL","ug/L",  "conc_corr_ugL + p_coprec_ugL; SRP without abiotic removal. NA for NH4-N",
  "flag_delta_ca_neg",   "-",     "TRUE when the reach gained calcium; model not parameterised for this direction"
)

write_csv(bind_rows(dict_btc_nutrients, dict_coprec),
          here("data_derived", "dictionary", "dict_btc_nutrients.csv"))
dict_coprec
```

## 9.6 Sensitivity of the coprecipitation correction to σ

`SIGMA_P` is the single most consequential constant in this correction —
`p_coprec_ugL` is exactly linear in it — and the published estimates
cluster into two physically distinct classes rather than forming one
continuous range: field/river waters (0.055-0.095 µmol/m², River Frome,
River Windrush, the Waterston chalk aquifer) and idealised synthetic
Ca(HCO₃)₂ solutions (0.13-0.22 µmol/m²), which run 2-4× higher because
they lack the competing solutes (Mg, Fe, dissolved organic matter) that
occupy sorption sites in real water. `SIGMA_P` adopts the low end of the
field cluster (River Frome, 5.5e-8 mol/m², following Cohen et al. 2013's
identical choice for a directly analogous karst stream); this section
carries the rest of that cluster and the laboratory ceiling forward as an
explicit envelope rather than asserting the point estimate as the only
possibility.

```{r coprec-sigma-sensitivity}
# p_coprec_molL was built as delta_ca_molL * (SIGMA_P * N_AVOG * D_CALCITE) * h_adsorbed,
# so for fixed delta_ca and h it is exactly proportional to SIGMA_P: rescaling
# the already-computed p_coprec_ugL is equivalent to, and cheaper than,
# rebuilding the join for each candidate sigma.
SIGMA_SENS <- c(
  field_low   = 5.5e-8,  # River Frome (House et al. 1986a) -- ADOPTED, = SIGMA_P
  field_high  = 9.5e-8,  # Waterston chalk aquifer (House et al. 1986a), field cluster upper end
  lab_ceiling = 1.5e-7   # synthetic Ca(HCO3)2 (House & Donaldson 1986); Cohen et al. (2013)'s
                         # laboratory upper-bound comparison value
)
stopifnot(isTRUE(all.equal(unname(SIGMA_SENS["field_low"]), SIGMA_P)))

coprec_sigma_sens <- btc_nutrients %>%
  filter(solute == "SRP", !is.na(p_coprec_ugL)) %>%
  select(event_id, sample_btc, p_coprec_ugL, conc_corr_ugL) %>%
  crossing(sigma_label = names(SIGMA_SENS)) %>%
  mutate(
    sigma_molm2       = SIGMA_SENS[sigma_label],
    p_coprec_ugL_sens = p_coprec_ugL * (sigma_molm2 / SIGMA_P),
    frac_of_srp_sens  = p_coprec_ugL_sens / conc_corr_ugL
  )

coprec_sigma_sens %>%
  group_by(event_id, sigma_label) %>%
  summarise(
    sigma_umolm2        = first(sigma_molm2) * 1e6,
    coprec_median_ugL   = median(p_coprec_ugL_sens, na.rm = TRUE),
    frac_of_srp_median  = median(frac_of_srp_sens,  na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(id_cols = event_id, names_from = sigma_label,
              values_from = c(coprec_median_ugL, frac_of_srp_median)) %>%
  arrange(event_id) %>%
  print(n = Inf)

write_csv(coprec_sigma_sens, here("data_derived", "coprec_sigma_sensitivity.csv"))

dict_coprec_sigma_sens <- tribble(
  ~column, ~unit, ~description,
  "sigma_label",       "-",      "field_low (adopted, River Frome) / field_high (Waterston aquifer) / lab_ceiling (synthetic solution)",
  "sigma_molm2",        "mol/m2", "SIGMA_P value used for this row, from SIGMA_SENS",
  "p_coprec_ugL_sens",  "ug/L",   "p_coprec_ugL rescaled to sigma_molm2; linear rescaling of the SIGMA_P-based value",
  "frac_of_srp_sens",   "-",      "p_coprec_ugL_sens / conc_corr_ugL, at this sigma"
)
write_csv(dict_coprec_sigma_sens,
          here("data_derived", "dictionary", "dict_coprec_sigma_sensitivity.csv"))
```

> **QAQC note.** This is the same check Cohen et al. (2013) ran on the
> Ichetucknee: moving from the field value to the laboratory ceiling
> should not be treated as a free parameter choice, because at the
> laboratory end `frac_of_srp` can approach or exceed unity — the point
> where the corrected series stops being a correction and becomes
> mostly model output (§9.5's QAQC note). Where `frac_of_srp_sens` stays
> small across the whole field-to-lab envelope, the choice of `SIGMA_P`
> cannot change a conclusion; where it diverges sharply between
> `field_low` and `lab_ceiling`, that divergence — not the point
> estimate — is what belongs in the manuscript, alongside the physical
> argument (§9.5) for why the field cluster is the appropriate class for
> a stream rather than a synthetic solution.

## 9.7 Event-level mass balance: does the correction respect the injected P mass?

`frac_of_srp` (§9.5) is a per-grab, concentration-domain ratio: at each
sampled instant, how large the correction is relative to the SRP
measured at that instant. The `median(frac_of_srp)` printed there is the
size of a *typical* correction, not the size of what the event actually
lost — it weighs every grab equally regardless of when it was taken or
how much water carried it, and `h_adsorbed` tracks phosphate speciation
(§9.2), so `p_coprec_ugL` rises and falls with SRP across the curve while
most events sample the tail more densely than the peak. A median across
grabs is therefore not a mass balance.

The mass-domain question — how much elemental P did coprecipitation
remove from the whole event, and is that plausible against the P
actually injected — needs a discharge-weighted integral over the sampled
curve, the same construction `vignettes/02_integration.Rmd`'s Section 6
already uses to get discharge from the NaCl mass balance (`hydro_probe`'s
`integral_Ct_mgLs`, `pracma::trapz`):

$$ m_{P,\mathrm{cop}} = Q \int_0^{T} C_{P,\mathrm{cop}}(t)\,dt
   \qquad
   m_{\mathrm{SRP}} = Q \int_0^{T} C_{\mathrm{corr}}(t)\,dt $$

with $Q$ the adopted event discharge (`discharge_Ls`, treated as steady
over the duration of a single slug event, matching that section's own
convention) and the integral taken by trapezoidal rule over the sampled
`time_since_release_s`. Two ratios follow:

$$ f_{\mathrm{mass}} = \frac{m_{P,\mathrm{cop}}}{m_{\mathrm{SRP}}}
   \qquad
   \%_{\mathrm{injectate}} = 100\,\frac{m_{P,\mathrm{cop}}}{m_{P,\mathrm{added}}} $$

`f_mass` is the mass-weighted analogue of `frac_of_srp` and is the
number to compare it against. `%_injectate` is an independent physical
check that does not use `conc_corr_ugL` at all — it asks what fraction
of the elemental P mass released at the addition point was removed by
coprecipitation before it passed the sampling station. `m_P,added`
converts the phosphate-salt mass on the field sheet (`added_mass_PO4_g`,
`molar_mass_P_nutrient`, `nutrient_addition_phosphate_data.csv`) to
elemental P using the same molar-mass ratio, not the salt mass itself.

`p_coprec_ugL` is integrated as measured, not clamped at zero: a
negative value is a real model output at `flag_delta_ca_neg` events (the
reach gained calcium — §9.5 already documents the model is not
parameterised for that direction), and clipping it would silently
overstate the correction there. `conc_corr_ugL` is clamped at zero in
its integral, matching the convention already used for `nacl_mgL_grab`
in `vignettes/02_integration.Rmd`'s Section 6 (a background-corrected
concentration can dip slightly negative from measurement noise around a
zero baseline; a transported mass cannot be physically negative).

```{r coprec-mass-balance}
p_added <- p_raw %>%
  distinct(stream, date, added_mass_PO4_g, molar_mass_P_nutrient)
stopifnot(!anyDuplicated(p_added[c("stream", "date")]))

coprec_mass_balance <- btc_nutrients %>%
  filter(solute == "SRP", !is.na(p_coprec_ugL)) %>%
  select(event_id, stream, date, time_since_release_s, conc_corr_ugL,
         p_coprec_ugL, flag_delta_ca_neg) %>%
  left_join(events %>% select(event_id, discharge_Ls, flag_discharge_invalid),
            by = "event_id") %>%
  left_join(p_added, by = c("stream", "date")) %>%
  arrange(event_id, time_since_release_s) %>%
  group_by(event_id) %>%
  summarise(
    n                      = n(),
    t_span_s               = diff(range(time_since_release_s, na.rm = TRUE)),
    discharge_Ls           = first(discharge_Ls),
    flag_discharge_invalid = first(flag_discharge_invalid),
    flag_delta_ca_neg      = first(flag_delta_ca_neg),
    # conc_corr_ugL can be NA on a grab even when p_coprec_ugL is
    # defined (a bad concentration reading does not stop the
    # geochemical calc, which does not use conc_corr_ugL at all).
    # A single NA in the window propagates through pmax()/trapz()
    # to NA for mass_srp_exported_ug and therefore frac_mass_coprec,
    # while mass_p_coprec_ug and pct_of_injectate_coprec (below) are
    # unaffected -- this flag names that asymmetry instead of
    # leaving it to be reverse-engineered from a bare NA.
    flag_conc_corr_na_in_window = anyNA(conc_corr_ugL),
    added_mass_PO4_g       = first(added_mass_PO4_g),
    molar_mass_P_nutrient  = first(molar_mass_P_nutrient),
    # discharge-weighted trapezoidal integral over the sampled curve; NA
    # where there are too few points to integrate or no valid discharge
    # (mirrors the guard used for integral_Ct_mgLs in 02_integration.Rmd).
    mass_srp_exported_ug = if (n >= 3 && t_span_s > 0 && !is.na(discharge_Ls))
                             discharge_Ls *
                               pracma::trapz(time_since_release_s, pmax(conc_corr_ugL, 0))
                           else NA_real_,
    mass_p_coprec_ug     = if (n >= 3 && t_span_s > 0 && !is.na(discharge_Ls))
                             discharge_Ls *
                               pracma::trapz(time_since_release_s, p_coprec_ugL)
                           else NA_real_,
    .groups = "drop"
  ) %>%
  mutate(
    frac_mass_coprec        = mass_p_coprec_ug / mass_srp_exported_ug,
    mass_P_added_g          = added_mass_PO4_g * (M_P / molar_mass_P_nutrient),
    pct_of_injectate_coprec = 100 * (mass_p_coprec_ug / 1e6) / mass_P_added_g,
    flag_insufficient_samples     = n < 3,
    flag_coprec_exceeds_injectate = !is.na(pct_of_injectate_coprec) &
                                     pct_of_injectate_coprec > 100
  ) %>%
  arrange(desc(frac_mass_coprec))

coprec_mass_balance %>%
  select(event_id, n, discharge_Ls, frac_mass_coprec, pct_of_injectate_coprec,
         flag_delta_ca_neg, flag_discharge_invalid, flag_insufficient_samples,
         flag_coprec_exceeds_injectate, flag_conc_corr_na_in_window) %>%
  print(n = Inf)

write_csv(coprec_mass_balance, here("data_derived", "coprec_mass_balance.csv"))

dict_coprec_mass_balance <- tribble(
  ~column, ~unit, ~description,
  "event_id",                     "-",     "Event key",
  "n",                             "-",     "Number of SRP grabs with a defined p_coprec_ugL",
  "t_span_s",                      "s",     "Time span covered by those grabs; integration domain",
  "discharge_Ls",                  "L/s",   "Adopted event discharge (events$discharge_Ls), treated as steady over the event",
  "mass_srp_exported_ug",          "ug",    "discharge_Ls * trapz(time_since_release_s, pmax(conc_corr_ugL, 0)); background-corrected SRP mass transported past the station over the sampled window",
  "mass_p_coprec_ug",              "ug",    "discharge_Ls * trapz(time_since_release_s, p_coprec_ugL); P mass removed by coprecipitation over the sampled window. Not clamped at zero -- see prose",
  "frac_mass_coprec",              "-",     "mass_p_coprec_ug / mass_srp_exported_ug; mass-weighted, event-integrated analogue of the per-grab median frac_of_srp in Sec. 9.5",
  "added_mass_PO4_g",              "g",     "Mass of the phosphate salt added, nutrient_addition_phosphate_data.csv",
  "molar_mass_P_nutrient",         "g/mol", "Molar mass of the salt added (type_P)",
  "mass_P_added_g",                "g",     "added_mass_PO4_g * (M_P / molar_mass_P_nutrient); elemental P mass injected",
  "pct_of_injectate_coprec",       "%",     "100 * (mass_p_coprec_ug/1e6) / mass_P_added_g; independent check, does not use conc_corr_ugL",
  "flag_delta_ca_neg",             "-",     "Carried from Sec. 9.5: TRUE when the reach gained calcium (model not parameterised for that direction)",
  "flag_discharge_invalid",        "-",     "Carried from events: TRUE when no valid discharge estimate exists for this event",
  "flag_insufficient_samples",     "-",     "TRUE when n < 3 (cannot integrate); mass columns are NA",
  "flag_coprec_exceeds_injectate", "-",     "TRUE when pct_of_injectate_coprec > 100 -- physically impossible, flags a remaining error rather than a real effect",
  "flag_conc_corr_na_in_window",   "-",     "TRUE when any conc_corr_ugL in the window is NA; frac_mass_coprec is then NA but pct_of_injectate_coprec (independent of conc_corr_ugL) is not -- expected, not a bug"
)

write_csv(dict_coprec_mass_balance,
          here("data_derived", "dictionary", "dict_coprec_mass_balance.csv"))

dict_coprec_mass_balance
```

> **QAQC note.** `frac_mass_coprec` is the number that should be compared
> to the per-grab median in §9.5, not treated as a replacement for it —
> a large gap between the two says the loss is concentrated in part of
> the curve the median under- or over-weights, which is itself worth
> reporting. `pct_of_injectate_coprec` is the stronger check: it is built
> from the addition record alone (`added_mass_PO4_g`,
> `molar_mass_P_nutrient`) and the coprecipitation model, with no
> dependence on the measured downstream SRP series, so it cannot inherit
> an error from `conc_corr_ugL`. Any `flag_coprec_exceeds_injectate ==
> TRUE` means the model is predicting more P removed than was released —
> physically impossible — and should be read as evidence of a remaining
> unit or join error, not as a real effect, per the same standard applied
> to `SIGMA_P` and the PHREEQC unit fixes earlier in this section.
>
> `flag_conc_corr_na_in_window` explains the one asymmetric case:
> `frac_mass_coprec` can be `NA` while `pct_of_injectate_coprec` is
> not, because a single `NA` in `conc_corr_ugL` propagates through
> the `mass_srp_exported_ug` integral but `mass_p_coprec_ug` never
> uses `conc_corr_ugL` at all. This is expected behaviour, not a
> second bug.
>
> **A note on circularity.** `p_coprec_ugL` — and therefore
> `conc_corr_coprec_ugL` — is a deterministic function of
> `delta_ca_mgL` (§9.2). Any uptake metric computed from the corrected
> series (Sw, Vf, areal uptake, or otherwise) therefore carries a
> component that is mechanically linear in ΔCa by construction,
> independent of any real biological or abiotic process. Comparing that
> metric, once corrected, back against ΔCa or calcium transfer and
> reporting agreement as confirmation of the coprecipitation mechanism
> is circular — the correction was built from ΔCa, so some relationship
> between the two is guaranteed rather than discovered. The evidence
> that is not circular is the relationship measured on the
> *uncorrected* series (nothing in `conc_corr_ugL` depends on ΔCa) and
> the independent mass-balance check above (`pct_of_injectate_coprec`,
> built from the addition record, not from ΔCa or the corrected
> concentration). A relationship that is absent uncorrected and appears
> only after applying the correction is particularly worth treating
> with suspicion for this reason before being reported as a finding —
> it is at least as consistent with the correction's own linear
> structure as with a newly-revealed process.
